\part{Philosophy of Computation}

\introquote{We're presently in the midst of a third intellectual revolution. The first came with Newton: the planets obey physical laws. The second came with Darwin: biology obeys genetic laws. In today’s third revolution, we're coming to realize that even minds and societies emerge from interacting laws that can be regarded as computations. Everything is a computation.}
{Rudy Rucker}

\textit{Computation} is an essential part of being human. Just as we act on our perceptions, feel our emotions, and daydream in our imaginations, we also compute with our \textit{reason} in order to find answers to the questions we have about life.

One can describe computation in a variety of ways. Some are poetic, others are more formal, and many we will discuss at length. In doing so, we will generally flow from poetry to formality, starting with broad statements and colorful examples and progressing toward a more sophisticated understanding. For now, we will say that computation is a process which resolves \textit{uncertainty}.

In his \textit{Theogony}, the Ancient Greek poet Hesiod (fl. 750 BCE) depicts a fascinating uncertainty called Chaos: a vast nothingness that preceded the creation of the Universe. From this void emerged the primordial deities, personifications of nature who gave life to the Titans and, by extension, to the Olympian gods. Curiously, the modern definition of chaos has nearly inverted. Now, one might say that a situation is chaotic if there is \textit{too much} going on. Despite their great conceptual difference, these meanings share a common property: they both evoke \textit{confusion}.

This place in which we find ourselves may indeed be post-Chaos, but confusion remains a familiar state for us. In fact, it is our natural state. We come into life confused about everything---crying, open-eyed in reaction to how much is going on around us. Slowly, we figure a few things out. Later, we consider many things \textit{certain}. As time ticks on and we carve our separate paths in the world, the wise ones among us come to realize the true depth of this confusion.

Although we are initially confused, we naturally pursue an understanding of the chaos that surrounds us. We perceive the goings-on of our environment and organize them into patterns that orient us and give meaning to the disorder of our existence. This is the crux of computation. It is the structured manipulation of \textit{data} into \textit{information} for the purpose of acquiring \textit{knowledge}. It shines a light into the darkness we call home.

These terms---\textit{data}, \textit{information}, and \textit{knowledge}---are broad enough to apply to a variety of systems. In fact, they appear in nearly every academic field and certainly within those that strive for objectivity. One can have data on, information about, and knowledge of anything that can be observed or experienced. As such, it is difficult to confine these terms to short and tidy definitions that are both rigorous and free of controversy. We will have to settle then for a long, pragmatic discussion.

Still, we must start somewhere. For example, you already have an idea of what information is, and it is likely that your intuition is broadly correct: information is a thing that tells you something. As we dive into more and more abstract topics, do not hesitate to, in this way, fall back on your existing vocabulary for support. A word used technically is more precise than the same word used casually, but the gist of it all is often the same. Or, at least, there is often some sense in which one reflects the other, like concepts linked in a good simile.

Recall the usual context in which technical language is born. A \textit{pattern} or \textit{phenomenon} is observed by a community, and it becomes the subject of casual discourse. It may even be a widespread concept that is discussed by society at large. Accordingly, the phenomenon is given an \textit{informal definition} and a \textit{name} (or, perhaps, many names). Interested theoreticians then formalize it; they describe its \textit{form} in unambiguous terms. They give it a \textit{formal definition} and bestow a more permanent name upon it, either one used previously or something new that, in the namers' opinion, better conveys the idea. In the latter case, the new term is disseminated throughout the community and perhaps to the wider public, where it may either replace or coexist with previous informal terms. Regardless, the namers must reach outside of the formalisms of their field to find a word from an informal, natural language that captures the essence of their abstract thought.

This is the beauty of a good piece of jargon: such a word will, to an amateur, shed a shallow but wide light into the vast space of an idea, but it will also illuminate the whole shebang in the eyes of an expert, evoking in its concise informality all that he or she has previously learned on the subject. In the spirit of this principle, our three words are defined softly below to act now as beacons in the dark of our ignorance and later as shining summaries of our hard-earned understanding:

\vspace{4mm}
\begin{displayquote}
    \textbf{Data} (sing. \textbf{datum}) are pieces of information that are meaningless when considered separately. They are observable lacks of uniformity in reality, each of which may be represented by a symbol that conveys its particular difference from the norm.

    \textbf{Information} is an ordered arrangement of data that has the potential to convey something meaningful. The data are arranged such that the information as a whole complies with the grammar of a language that can, in theory, be understood.

    \textbf{Knowledge} is an understanding of information that emerges given proper interpretation and purposeful thought. It is usually associated with truth and may be considered a body of true beliefs.
\end{displayquote}
\vspace{4mm}

The journey to an honest appreciation of these terms will be an arduous one, and subtly so, as it will require an open mind and an ample store of time set aside for the contemplation. Nevertheless, walking that inner path is unquestionably worth the cost when one considers how pervasive these concepts truly are. They can describe any event, including those that occur in computers as they run their software and in brains as they continuously perceive and conceive of their own respective conscious experiences. This chapter explores Man's longstanding interest in these scraps of reality: our inquiry into the smallest details of Nature; our concretization of abstract thought through language, signage, and code; our distillation of human reason into mechanical rulesets.

There is a fourth word---\textit{wisdom}---but it is even harder to pin down than the previous three. It is a lofty word---so lofty that it is difficult to say anything about it without sounding like a proselytizer. Its utterance conjures up images of ashen-haired, big-bearded men in togas or robes, speaking of grandiose things or passing time in peaceful admiration of the world. It has no formal definition because its form is unknown, close enough to touch momentarily but always out of reach. Furthermore, it is bound to messy human concerns like emotions, morals, and purpose. How, then, could wisdom be relevant to the study of something as well-defined as computation?

\input{\volOne/c1-cognitive-ontology}
\input{\volOne/c2-metaphysics-natural-philosophy-and-physics}
\input{\volOne/c3-information-and-communication}
\input{\volOne/c4-logic-and-mathematics}

% Dinkus

% Conclusion to Volume I
What is the strange loop? What is wisdom?

