\part{Theory of Computation}

\introquote{Theory is relevant to you because it shows you a new, simpler, and more elegant side of computers, which we normally consider to be complicated machines. The best computer designs and applications are conceived with elegance in mind. A theoretical course can heighten your aesthetic sense and help you build more beautiful systems.}
{Michael Sipser}

The \textit{theory of computation} is a field of study that is related to both mathematics and computer science. It is concerned with a process known as \textit{computation}, which refers to any kind of \textit{calculation} (the transformation of one or more inputs into one or more outputs) that is accomplished by following an \textit{algorithm} (a set of rigorously defined steps). It is also concerned with the formalization of theoretical models of computation known as \textit{abstract machines} or \textit{abstract automata}. Also of interest are the \textit{computational problems} that can be solved by these abstract machines and how efficiently they can be solved. The theory of computation gives us the formal definition of a computer and the scope of what a computer can accomplish.

Much of the work in the theory of computation is$\dots$ quite abstract. While calculation and computation have been practiced by humans for thousands of years, the concept was only formalized by mathematicians, logicians, philosophers, linguists, and computer scientists in the past 150 years. This part of the guide will delve into some fairly academic topics with a decent amount of mathematical rigor, but it is written to be accessible and entertaining.

The field has three branches: \textit{automata theory}, \textit{computability theory}, and \textit{computational complexity theory}, each of which asks fundamental questions pertaining to how mathematical calculations can be automated, which kinds of problems can be solved by a machine, and how many resources are required to solve a particular problem. We will touch on the major discoveries made in each of these branches and discuss why they are significant to computer science and programming in general.

\input{\volTwo/c5-automata-theory}
\input{\volTwo/c6-computability-theory}
\input{\volTwo/c7-computational-complexity-theory}
\input{\volTwo/c8-computer-architecture}

% Dinkus

% Conclusion to Volume II

