%--------------------------------------------------------------------------------
%--------------------------------------------------------------------------------
%    CHAPTER I: COMPUTATION
%--------------------------------------------------------------------------------
%--------------------------------------------------------------------------------

\part*{Philosophy of Computation}
\addcontentsline{toc}{part}{\tocpartglyph Philosophy of Computation}

\vspace{4mm}
\begin{displayquote}
	\textit{We're presently in the midst of a third intellectual revolution. The first came with Newton: the planets obey physical laws. The second came with Darwin: biology obeys genetic laws. In today’s third revolution, we're coming to realize that even minds and societies emerge from interacting laws that can be regarded as computations. \textbf{Everything is a computation.}}
	\begin{flushright}
		---Rudy Rucker
	\end{flushright}
\end{displayquote}
\vspace{4mm}

\textit{Computation} is an essential part of being human. Just as we act on our perceptions, feel our emotions, and daydream in our imaginations, we also compute with our \textit{reasoning} in order to find answers to the questions we have about life. \\

One can describe computation in a variety of ways. Some are poetic, others are more formal, and many we will discuss at length. In doing so, we will generally flow from poetry to formality, starting with broad statements and colorful examples and progressing toward a formal understanding. For now, we will say that computation is a process that resolves \textit{uncertainty}. \\

In his \textit{Theogony}, the Ancient Greek poet Hesiod (fl. 750 BCE) depicts a fascinating uncertainty known as Chaos: a vast nothingness that preceded the creation of the Universe. From this void emerged the primordial deities, personifications of nature who gave life to the Titans and, by extension, to the Olympian gods. Curiously, the modern definition of \textit{chaos} has nearly inverted. Now, one might say that a situation is \textit{chaotic} if there is \textit{too much} going on. Despite their great conceptual difference, these meanings share a common property: they both evoke \textit{confusion}. \\

This place in which we find ourselves may indeed be post-Chaos, but confusion remains a familiar state for us. In fact, it is our natural state. We come into life confused about everything---crying, open-eyed in reaction to how much is going on around us. Slowly, we figure a few things out. Later, we consider many things \textit{certain}. As time ticks on and we carve our separate paths in the world, the wise ones among us come to realize the true depth of their confusion. \\

Although we are initially confused, we naturally pursue an understanding of the chaos that surrounds us. We perceive the goings-on of our environment and draw \textit{information} from them, piecing together a \textit{knowledge} that orients us and gives meaning to the disorder of our existence. This is the crux of computation. It is the structured manipulation of \textit{data} into \textit{information} for the purpose of acquiring \textit{knowledge}. It shines a light into the darkness we call home. \\

These terms---data, information, and knowledge---are broad enough to apply to a variety of systems. In fact, they appear in nearly every academic field and certainly within those that strive for objectivity. One can have data on, information about, and knowledge of anything that can be observed or experienced. As such, it is difficult to confine these terms to short and tidy definitions that are both rigorous and free of controversy. We will have to settle then for long, pragmatic definitions. \\

In this chapter, we will explore what computation is and why it matters. In some sense, computation is a kind of game in which \textit{pieces} are moved and transformed within a \textit{game space} according to a predetermined set of \textit{rules}. However, unlike most games, computation has no set goal. Nature shuffles its hazy pieces about, following arbitrary rules for reasons we cannot ascertain. Man, on the other hand, explicitly defines its pieces and rules, modeling them after Nature's strange mechanics. We invent \textit{win conditions} and play toward them, hoping that our computations will result in general solutions that will elucidate the knowledge we seek. \\

\begin{description}
	\item[Section 1:] \textit{Information and Communication}, covers the pieces \dots
	\item[Section 2:] \textit{Reasoning and Logic}, covers the rules \dots \\
\end{description}

Here, we learn how this game is played and how it is woven into the fabric of our reality. In the chapter following this one, we model the game of computation mathematically and find the theoretical limits of what it can tell us. Beyond that, we learn how to be an excellent player, covering \textit{techniques} for efficient problem solving and \textit{tools} for designing beautiful, maintainable systems.

%--------------------------------------------------------------------------------
%    SECTION: INFORMATION AND COMMUNICATION
%--------------------------------------------------------------------------------

\toclineskip
\section{Information and Communication}

\vspace{4mm}
\begin{displayquote}
	\textit{For Wiener, entropy was a measure of disorder; for Shannon, of uncertainty. Fundamentally, as they were realizing, these were the same.}
	\begin{flushright}
		---James Gleick
	\end{flushright}
\end{displayquote}
\vspace{4mm}

The most general concepts are among the hardest to grasp, and \textit{information} is about as general as you can get. Potentially \textit{everything} can be described in terms of information---the physical world, all of literature and written language, and perhaps even our conscious experience. Thus, our path to a proper understanding of information and how it relates to reality will be an arduous one, requiring an open mind and an ample store of time set aside for contemplation. But it will also be an elegant one, weaving its way through philosophy and physics and arriving at an \textit{information theory} that we can use to explain the behavior of engineered systems, human language, and the intersection of the two: mechanical computation. \\

You already have an idea of what information is, and it is likely that your intuition is broadly correct: information is something factual that tells you something. As we dive into more and more abstract topics, do not hesitate to, in this way, fall back on your existing vocabulary for support. A word used technically is more formal and precise than the same word used casually, but the gist of it all is often the same. \\

Recall the usual context in which technical language is born. A phenomenon is observed by a community, and it becomes the subject of casual discourse. It may even be a widespread concept that is discussed by society at large, in some way or another. The phenomenon is accordingly given an \textit{informal definition} and a name (or, perhaps, many names). Interested theoreticians then study the structure of the phenomenon and \textit{formalize} it. They give it a \textit{formal definition} and bestow it a more permanent name, either one used previously or something new that is, in the namers' opinion, more apt. In the latter case, the new term is disseminated throughout the community and perhaps to the wider public, where it may either replace or coexist with previous informal terms. \\

Regardless, the namers must reach outside of the formalisms they have constructed and find a word from an informal language that captures the essence of their abstract thought. This is the beauty of good jargon: such a word will, to an amateur, shed a shallow but wide light into the vast space of an idea, but it will also illuminate the whole shebang in the eyes of an expert, evoking in its concise informality all that he or she has previously learned on the subject. \\

In the spirit of this principle, four words are given below, defined softly, that will act now as beacons in the dark of our ignorance and later as summaries of our hard-earned understanding: \\

\begin{displayquote}
	\textit{
		\textbf{Data} are pieces of information that are meaningless when considered separately. They are observable lacks of uniformity in reality, each of which may be represented by a symbol that conveys its particular difference from the norm. \\~\\
		\textbf{Information} is an ordered arrangement of data that has meaning. The data are arranged such that the information as a whole complies with the grammar of a language that can, in theory, be understood. \\~\\
		\textbf{Knowledge} is an understanding of information that emerges given proper interpretation and thought. It is usually associated with Truth and may be considered a body of true beliefs. \\~\\
		\textbf{Wisdom} is the capacity to think and act in a way that is right. It involves the use of knowledge, a prudent choice of means and ends, and a healthy doubt of one's own understanding. It is born of an acknowledgment of how little one truly knows and how little control one truly has. \\
	}
\end{displayquote}

With these fundamental terms in mind, let us take a page from Hesiod and begin our tale with an uncertain chaos. Let us regard reality as a baby might: confusing but ultimately intelligible through \textit{categorization}.

\subsection{Categories of Being}

% Describe chaos, pure perception with no thought, no categories, just raw reality. Quality. Qualia (what is it like to...). Experience. Truth.
% Chaos is phenomena. Just pure sensory experience, no analysis.
% First thoughts

% The Universe is a jumble of data, absorbed by our various sensors---eyes, ears, and the rest---and organized by our brains into meaningful information. We then interpret this information and come to understand its meaning. And in doing so, we acquire a personal conclusion that may or may not align with conclusions made by other observers. Hence comes the motivation for logical debate: to determine which conclusions are \textit{true}. \\

% A category system is an interpretation of reality, a firm partitioning of some phenomena from others.

% What is categorization? A mental process?
% What is a category? Philosophy version. Give the math version at the end of the metaphysics section? Because category theory is about studying things relationally and classical metaphysics was more object-oriented.
% What is a category of being? An ontological category, a category for things that exist.

% Important: is there a most general category?
% If so, thing\entity\individual is the most general category (provides no information). And an object is any thing that is suitable for thought ("thing presented to the mind"). Are all things objects? Who the fuck knows. Let's assume yes.

% Categorizing in other fields: history periods, biology taxa

% Read all of Categorization
% Read all of Ontology (information science)

% In order to clear the chaos, we must first ask what there is. What can we say about the things that exist? Can we fundamentally distinguish some from others, or is it all just a big, soupy mess? Furthermore, if we indeed say that some things are $A$ and the rest are $B$, can we also claim that such a distinction is \textit{actually} present in reality or are things $A$ and $B$ only because that is how our minds perceive them? These are some of the oldest questions in \textit{metaphysics}, the study of what is. \\

% What is a metaphysical object? It's like a trope. It's a pattern that is recognizable, and it comes up frequently enough that there is a name for it. It's a description of some thing that often appears in reality. It is a \textit{type} of thing. We will discuss many of these metaphysical objects in the chapter Types and Structures.

% Eventually, a baby is capable of thinking symbolically (language)
% Syntax, semantics, pragmatics
% Sorting something alphabetically vs. categorizing semantically

% Does category in philosophy mesh with category in math?
% In category theory, a category is a class of objects and a class of relations (morphisms). Individuals and classes have attributes, but individuals are not usually considered in category theory (considering that categories have classes which are *collections*).
% So a category in math is a collection of objects + their properties + their relations?

% We will cover the main categories of being generally and will also give some of the most important examples.

\subsubsection{Properties}

% Bundle theory, bare particularism.
% has-a relation

A common approach to this challenge is creating a system of \textit{categories} that partition reality at the highest level. For example, one could suggest that there is a fundamental difference between a \textit{quantity} and a \textit{quality}. If these are considered distinct categories, we can classify \textit{measurable} or \textit{countable} stuff like \textit{length}, \textit{mass}, \textit{number}, and \textit{monetary value} as quantitative and \textit{experiential} stuff like \textit{softness}, \textit{roundness}, \textit{flavor}, and \textit{beauty} as qualitative. \\

The quantity-quality distinction is a generally accepted one. However, the dichotomy is not so absolute. For example, softness is listed above as a quality. We might expect, then, that hardness is a quality. For example, rocks are hard and puppy ears are soft. But hardness is also a quantitative measurement of resistance to plastic deformation. Thus, the word \textit{hardness} can refer to both quantitative and qualitative phenomena. The categorical difference between these hardnesses, which exists independent of language, is identified through the context in which the word is used. A hardness can be measured with numbers, but it can also simply be \textit{experienced} via touch. \\

% All quantities are also qualities. It would be hard to argue that all qualities are also quantities in any meaningful way. For example, using millihelens as a quantitative unit of beauty. It doesn't really work.

%----------------------------------------

\textbf{Space and time} are two features of our reality. As they appear to us, the former is an expanse in three dimensions, and the latter can be thought of as an arrow that extends steadily in one dimension, from past to future. They are inextricably woven into the fabric of our experience, and yet their true nature eludes us. \\

Consider a space that contains many particles of \textit{substance}, all of which move about constantly as time goes on. Suppose that, one by one and randomly, these particles blink out of existence forever. When only one particle remains, is that which is \textit{not} the particle still considered space? Further, can we still say that this particle moves? It remains surrounded by a void, interacting and relating with nothing at all, participating in no distinguishable events. And if indeed nothing changes, does time still tick? If so, will space and time \textit{mean} anything after that last, lonely bit of matter disappears and everything becomes a uniform non-existence? \\

These are the questions that pertain to whether space and time are \textit{absolute} or \textit{relative}. If space is absolute, like a fixed three-dimensional grid that pervades the material realm, we might ask where the origin $(0,0,0)$ is. If such a location were to exist, every object and particle in the Universe would have an absolute position in three coordinates. In other words, position would be a distinguishing feature of each and every object, and it would not be defined in relation to other objects, but to this primordial \textit{center of the Universe}. \\

If space is instead relative, a conceptual origin can be placed wherever one likes. Distances can then be measured from this point. Of course, this can be done in an absolute space as well. Thus, the absolutist viewpoint commits the cardinal sins of being \textit{unobservable} and being \textit{useless}. To any human observer, a reality with absolute space would look no different than a reality with relative space---objects would not change in any way after being related to an arbitrary center. Access to these absolute positions also would not provide us with any additional information---they would just be fixed offsets of any positions measured from a relative point. \\

Perhaps then, space is is better thought of as \textit{relational} rather than \textit{objective}. Perhaps it is not a hidden grid-like object itself, but an \textit{ordering} that is composed of relations between objects. And perhaps then it is also better thought of as a product of the \textit{mind} rather than as a physical thing. That is, space may instead be an \textit{abstraction} that our minds make by estimating the relative distance between two objects and comparing it to every other potential spatial relation in our surroundings. In this view, space is simply a web of relations that is constructed by a sentient being. \\

Similarly, time may not be a hidden clock-like object, but instead a series of relative temporal intervals between \textit{events}. And if we only know that time passes because of the changes we perceive in space, it is possible that time is an abstraction as well, an ordering of events made by a sufficiently conscious mind. Furthermore, if space and time are both relational and coupled, perhaps it is better to think of them as a unified \textit{spacetime} despite our predispositions to think of them separately. \\

The philosophy of space and time is full of eternal questions about infinity, substance, void, uniformity, asymmetry, and consciousness. Comparatively, \textit{computational space and time} are simple. \\

% CS is a closed-world.
% Time in computer science is often based on the changing of state.

\subsubsection{Objects}

% Discuss how object and subject switched.

% Objects vs. events. Is there a difference? Maybe not, but there at least appears to be one. Humans have traditionally discerned a difference, notably in language. Perhaps the difference is only conceptually convenient. Is an apple distinct from its decaying over time? Is computer state really a meaningful object with clock speeds in the gigahertz? Is a thunderstorm an event, or just a system of many, tiny objects with different properties, all changing with time?

This curiosity aside, we could also design our category system such that quantities and qualities are both considered \textit{properties} of some very general sort of thing. Philosophers have long considered the idea of a highest-level category and have ascribed a variety of names to its members: \textit{entity}, \textit{being}, \textit{thing}, \textit{term}, \textit{individual}, etc. Each carries its own subtle connotations. We will use the name \textit{object} because it is the default choice in current philosophical practice and because it has bled into the technical language of computer science with its meaning largely unchanged. \\

How general is this category of objects? Namely, are properties objects? If they are, then our system collapses to a single category, which really means that no distinctions about reality can be made at all. The category of objects thus becomes the soupy mess we are trying to escape. For now, let us assume that properties are \textit{not} objects. Instead, they are something else, and they \textit{relate} to objects in some way, giving them their characteristics. From these relations, the diversity of existence unfolds. \\

%----------------------------------------

The \textbf{abstract-concrete dichotomy} classifies \textit{things} as being either \textit{abstract} or \textit{concrete}. Despite the fact that most people recognize a difference between an abstract and a concrete thing, this dichotomy does not have a universally accepted definition. That said, being abstract is often defined as being both \textit{non-spatiotemporal} and \textit{causally inefficacious}. That is, something is abstract if and only if 

\begin{enumerate}
	\item it does not exist in \textit{space} or in \textit{time} (or, alternatively, in \textit{spacetime}), and
	\item it cannot be the \textit{cause} of any \textit{effect}. \\
\end{enumerate}

However, one might argue that nothing is truly causally inefficacious. Indeed, if something was truly unable to cause \textit{anything}, it would not be able to affect the \textit{thoughts} of any philosopher, mathematician, or scientist for whom such an abstract thing would be of interest. And yet \textit{mathematical objects} (e.g. a number, a line, a cube, the sine function, etc.) are often considered abstract, despite the fact that they affect the thoughts of mathematicians daily, thoughts which in turn prompt concrete behavior, such as drawing diagrams. Perhaps, causal inefficacy is too strict a requirement. \\

Let us consider why one might want to refer to something that is abstract. An abstract object references \textit{properties} that many concrete, causally efficacious objects have. For example, a sphere does not exist in space or time; it is \textit{nowhere} and \textit{never}. It also does not seem capable of causing an effect in the way that, say, a ping-pong ball can. And while the abstract sphere may not have the same sort of causal efficacy as concrete spherical objects, it does have a \textit{relation} to those objects. That is, the concept of sphericality is \textit{related} to the \textit{class} of concrete objects that may be reasonably modeled as spheres. Thus, an abstract object is, at least, convenient because it can identify a class of concrete objects that is of interest by specifying relevant properties. \\

In light of recent advancements in fields like neuroscience and artificial intelligence, many believe that the human mind is nothing more than a physical system, albeit a particularly complicated one, that can be modeled mathematically like any other. So it may be that abstract objects are just interpretations of concrete electrochemical signals in our brains. Regardless,  \\

% "Abstraction takes us further from reality but casts a wider light on it."
% Abstraction allows you to handle large quantities of information as single units. This allows you to manipulate information more efficiently (say more, do more).

% Are there abstract things at all (nominalism)? Is chess an abstract object? Is a novel an abstract object?
% Causality.
% Concrete objects have substance. An object of spacetime is an event.
% Is a set of books abstract or concrete?

\subsubsection{Classes}

% Extension-intension
% Universals and particulars
% Individuals (urelements) and sets (individuals are deemed redundant in modern set theory). Pure and impure sets.
% Subset, superset, class.

A \textbf{domain of discourse} is a set of all the things under discussion. It is also called a \textit{universe}, particularly in the field of mathematical logic. It specifies, out of all conceivable objects of study, those objects which are pertinent to the matter at hand. It gives \textit{context} to what we say. \\

A \textit{discourse} is a conversation, and its \textit{domain} is the subject of the conversation. A domain may also refer to an entire field of study, perhaps one whose conversation has been developing for thousands of years (e.g. physics, whose domain is physical reality). Each field of study has an accompanying domain of discourse, the set of objects within its purview. Working in such a field is thus akin to exploring a \textit{universe} populated with such objects. \\

As we observe the objects of a universe, we can make \textit{logical statements} about them. We can also ensure our statements are consistent by using a \textit{formal system of deduction} to prove them from a set of \textit{axioms}. And if we compile every statement that can be proven in this system---the axioms and their consequences---we will form a \textit{theory}, a set of true statements (or \textit{theorems}) that describes the universe in question. \\

The word \textit{theory} may also refer to a \textit{theory-in-progress}, a body of logical work to which theoreticians contribute. We might consider this an \textit{open theory}, a theory for which we do not have all possible knowledge. In contrast, a \textit{closed theory} is one for which we know everything there is to know. \\

Similarly, we can describe universes as open or closed. The \textit{open-world assumption} is made when we do not have complete knowledge of a system: if a statement is not known to be true (i.e. it is not a member of our theory), we do not assume anything about its truth value. This is the case for any system that is \textit{discovered}, like the various systems of nature. \\

Computational systems, however, are not discovered but \textit{designed}. Under normal conditions, it is possible to know everything that happens during a digital computation---all of the information appears in a discrete, finite space of computer memory. Thus, we are omniscient with regard to this universe and should make the \textit{closed-world assumption}: if a statement is not known to be true, it must be false. \\

%----------------------------------------

The \textbf{type-token distinction} \\

% A natural kind is a collection of tokens that have something in common.
% Extensional and intentional definitions of class.

\begin{tcolorbox}[breakable, enhanced, colback=textbook-blue, sharp corners]
	\vspace{3mm}
	\begin{center}
		\textbf{The Universal and Particular Notions of Chess}
	\end{center}
	% Chess example, physical boards and pieces, game states, the abstract idea of chess. "When was chess invented?". An instance of a chess game vs. all actual chess games vs. all possible chess games vs. the class of chess.
	
	% Has the abstract idea of chess existed for eternity? Or was it invented? If there is no sentient life, are there spheres?
	% Are mathematical objects mind-dependent? Could a different kind of sentient mind recieve the same sensory data and interpret it with an alien sort of ontology?
	
	As an illustrative example, consider all of the games of chess that are being played right now, currently progressing or left half-finished and awaiting the next move. Consider also all of the \textit{historical} games of chess that have been played to completion or total abandonment since the invention of chess.
	\vspace{3mm}
\end{tcolorbox}
\vspace{2\baselineskip}

\subsubsection{Relations}

\textbf{Identity} seems to be a mandatory quirk of being. It is difficult to imagine an existence in which things are not themselves. And yet, it is also difficult to define, in any meaningful way, this supposedly crucial concept. \\

% Identity is tricky in philosophy but easy in math.
% Sorites paradox, Theseus paradox

%----------------------------------------

The \textbf{map-territory relation} \\

% Symbols represent objects, but they are not the objects.
% Models represent reality, but they are not reality.
% Exoteric and esoteric.

% The syntax, semantics, and pragmatics of a language (relating symbols to concepts, and doing so sensitive to context)

% Syntax includes grammar, phonology, and orthography.
% Truth-falsity.
% Programming is the conversion of natural language semantics to formal semantics written in formal syntax.
% A high-level language just has a formal syntax that is closer to natural syntax.

\subsubsection{Taxonomies}

% Subsumation - taxonomies
% Instantiation
% Aggregation, composition, containment
% Type system, ontology

Say you want to \textit{communicate} an idea $X$ to someone. That is, you would like to convey some \textit{information} to them so that you might together establish a shared body of \textit{knowledge}. $X$ is on the tip of your tongue, but you cannot quite put it into words. Accordingly, you express a related idea $Y_0$, hoping that you might be able to suss out your target $X$ with the help of your fellow interlocutor. This proxy idea is either too specific or too general, a \textit{subclass} or \textit{superclass} of the \textit{class} $X$, and you must replace it with another idea, preferably $X$ or at least some $Y_1$ that is closer to $X$ than $Y_0$. And so on and so forth, ideas are proposed until some candidate $Y_n=X$ is found. The conversation could go as follows: \\

\begin{addmargin}[8em]{2em}
\begin{flushright}
"Dude, what's that thing with the moving parts that beeps sometimes? And it, like, gets hot if it moves too fast? \\[1mm]

\mydots \\[1mm]

It's like a \textit{Machine} of some sort, but more \textit{specific}." \\[\baselineskip]
\end{flushright}
\end{addmargin}

\begin{addmargin}[2em]{8em}
"Oh yeah, those expensive things with the metal chassis and all the buttons? About 15 feet long with two rows of seats? \\[2.25mm]
Sounds like a good ol' fashioned \textit{Sedan}, my dude." \\
\end{addmargin}

\begin{addmargin}[8em]{2em}
	\begin{flushright}
		"Nah, it's not a sedan. That's close, but I'm thinking of something more \textit{general} than that. It's definitely a \textit{type} of \textit{Vehicle} though." \\[\baselineskip]
	\end{flushright}
\end{addmargin}

\begin{addmargin}[2em]{8em}
	"Well, there are a lot of \textit{Vehicle} types out there, but I'm guessing this has to do with a \textit{Car}, specifically my---" \\
\end{addmargin}

\begin{addmargin}[8em]{2em}
	\begin{flushright}
		"Of course! I was thinking of a \textit{Car}. \\[2mm]
		Ok, now that we're on the same page, about \textit{your car}\dots \\
		My bad, man. I crashed it into the lake last night." \\[\baselineskip]
	\end{flushright}
\end{addmargin}

\begin{addmargin}[2em]{8em}
	\mydots \\
\end{addmargin}

\begin{addmargin}[8em]{2em}
	\begin{flushright}
		"Hey, if it makes you feel better, Joey dared me to do it, so there wasn't really an alternative." \\[\baselineskip]
	\end{flushright}
\end{addmargin}

In this scenario, Interlocutor $\mathcal{A}$ introduces the problem: there is \textit{information} to \textit{communicate}, and it defines a \textit{class} $X$. And a \textit{class}, we understand, is simultaneously a collection of individual \textit{objects} and an \textit{abstract object} itself, a formal definition of the collection to which all members comply. $\mathcal{A}$ grasps for $X$, throwing out descriptions to lure it closer, but it evades. $\mathcal{A}$ then takes the class $Y_0$, \textit{Machine}, as a starting point and states that it is a \textit{superclass} of $X$. \\

Interlocutor $\mathcal{B}$ responds with \textit{specialization}, adding assumed details to the definition of $Y_0$. Thus, $Y_1$ is proposed: the class \textit{Sedan}. Luckily, the assumptions are not totally off-base; $Y_1$ turns out to be a \textit{subclass} of $X$. $\mathcal{A}$ confirms this and so returns with a \textit{generalization} that conjures $Y_2$, the class \textit{Vehicle}. $\mathcal{B}$ \textit{specializes} again, offering up as $Y_3$ a subclass of \textit{Vehicle} called \textit{Car}. Finally, $\mathcal{A}$ accepts this class as the target $X$ and ends the search. And with this mutual knowledge of \textit{Car} established, $\mathcal{A}$ then refers to an \textit{instance} of \textit{Car}, a particular car $b$ that belongs to $\mathcal{B}$. And with both interlocutors now cognizant of exactly what is being discussed, $\mathcal{A}$ swiftly informs $\mathcal{B}$ that his ride is totaled.

% Taxonomy diagram
\begin{center}
	\resizebox{\textwidth}{!}{
		\begin{tikzpicture}[scale=0.2]
		
			%--- GRAY TREES -------------------------
		
		    \foreach \x in {0,...,3}
		        \foreach \y in {1,...,5}{
		            
		            % Big dashed lines
		            \draw [gray,dashed,thick] (\x*17,0) -- ++(90+\y*15:8);
		            \draw [gray,dashed,thick] (\x*17,0) -- ++(180+\y*15:8);
		            
		            % Small dashed lines
		            \draw [gray,dashed,thick] (\x*17,0) ++(90+\y*15:8) -- ++(90+\y*15-15:3);
		            \draw [gray,dashed,thick] (\x*17,0) ++(90+\y*15:8) -- ++(90+\y*15+15:3);
		            \draw [gray,dashed,thick] (\x*17,0) ++(180+\y*15:8) -- ++(180+\y*15-15:3);
		            \draw [gray,dashed,thick] (\x*17,0) ++(180+\y*15:8) -- ++(180+\y*15+15:3);
		            
		            % Small circles
		            \draw [gray,thick,fill=white] (\x*17,0) ++(90+\y*15:8) circle (0.55);
		            \draw [gray,thick,fill=white] (\x*17,0) ++(180+\y*15:8) circle (0.55);
			    }
		    
		    % Extra gray tree for Y_1 at 180 degrees
		    \draw [gray,dashed,thick] (0,0) -- ++(180:8);
		    \draw [gray,dashed,thick] (0,0) ++(180:8) -- ++(180-15:3);
		    \draw [gray,dashed,thick] (0,0) ++(180:8) -- ++(180+15:3);
		    \draw [gray,thick,fill=white] (0,0) ++(180:8) circle (0.55);
		    
		    %--- CURVES AND BIG ARROWS --------------
		    
		    % Curve coordinates
		    
		    % Red
		    \coordinate (R1) at ($(17,0)+(-65:4)$);
		    \coordinate (R2) at (24,-12);
		    \coordinate (R3) at (20,-22);
		    
		    % Blue
		    \coordinate (M1) at (8,-18);
		    \coordinate (M2) at (-4,-14);
		    \coordinate (M3) at (5,-1);
		    
		    % Green
		    \coordinate (L1) at (6,-8);
		    \coordinate (L2) at (7,-15);
		    \coordinate (L3) at (-5.5,-22);
		    \coordinate (END) at ($(-12,-10)+(-80:2)$);
		    
		    
		    % MIDDLE curve
		    \draw [thick] plot [smooth]
		    (M1) .. controls (M2) and (M3) .. (L1);
		    
		    % Big arrows
		    \draw [thick, rounded corners, fill=white] (-6.5,13) -- ++(0,4) -- ++(50,0) -- ++(0,2) -- ++(9,-4) -- ++(-9,-4) -- ++(0,2) -- cycle;
		    \node at (10.5,15) {\Large \textsc{Generalization}};
		    
		    \draw [thick, rounded corners, fill=white] (54.5,-13) -- ++(0,-4) -- ++(-50,0) -- ++(0,-2) -- ++(-9,4) -- ++(9,4) -- ++(0,-2) -- cycle;
		    \node at (37.5,-15) {\Large \textsc{Specialization}};
		    
		    % RIGHT curve
		    \draw [thick] plot [smooth]
		    (R1) .. controls (R2) and (R3) .. (M1);
		    
		    % LEFT curve
		    \draw [-{Latex[length=5mm,width=2mm,angle=60:6pt]},thick] plot [smooth]
		    (L1) .. controls (L2) and (L3) .. (END);
		    
		    %--- NODES AND SMALL ARROWS -------------
		    
		    % Arrows
		    \foreach \x in {1,2,3}{
		    	\draw [-{Latex[length=5mm,width=2mm,angle=60:6pt]},thick] (17*\x,0) -- +(-13,0);
		    }
	    	\draw [-{Latex[length=5mm,width=2mm,angle=60:6pt]},thick,dashed] (17*3+8,0) -- +(-4,0);
		    
		    % Y nodes
		    \draw [thick,fill=white] (0,0) circle (4) node {\Large $Y_1$};
		    \draw [thick,fill=textbook-blue] (17,0) circle (4) node {\Large $Y_3$};
		    \draw [thick,fill=white] (2*17,0) circle (4) node {\Large $Y_2$};
		    \draw [thick,fill=white] (3*17,0) circle (4) node {\Large $Y_0$};
		    
		    % b node
		    \draw [thick,fill=textbook-blue] (-12,-10) circle (2);
		    \node at (-12,-10) {\Large $b$};
		    
		    % Bezier curve assistant (used for sculpting curves)
		    % \draw [thick,red] (R1) -- (R2) -- (R3) -- (M1);
		    % \draw [thick,blue] (M1) -- (M2) -- (M3) -- (L1);
		    % \draw [thick,green] (L1) -- (L2) -- (L3) -- (END);
		    
		\end{tikzpicture}
	}
\end{center}

% Diagram -> Generalization/Specialization -> Top and Bottom/Type Theory -> Generalization is theory building, specialization is theory using -> information is both fundamental and very general -> because it is so general, we need to pull from a lot of fields to properly characterize it.

In the above diagram, we see the $Y$ classes and their relations to each other. The arrows between them are \textit{subsumption} relations. They point from a \textit{parent} class $P$ to a \textit{child} class $C$, and we say that $P$ \textit{subsumes} $C$ (i.e. it encompasses the entirety of $C$). In set theory notation, $P\subset C$. Alternatively, we can reverse the arrows and call them instead \textit{is-a} relations, which point from $C$ to $P$. Every $c\in C$ is a member of the more general class $P$, and thus every $c$ \textbf{is a} $P$ of some \textit{type}. \\

Of course, any given type $T$ can have many \textit{subtypes}, and those subtypes can have their own subtypes. A subtype of $T$ builds on the information that defines $T$, adding new information that specializes the underlying parent class. And with enough specialization, these subtypes may become quite unlike each other at a glance, sharing little else than a distant relation to $T$. This branching nature of subsumption is depicted above in gray. \\

Like branches that furcate from an old oak tree and fork into smaller branches toward the leaves, types spread in a \textit{recursive} manner, each one potentially a parent of arbitrarily many children and an ancestor of arbitrarily long lineages. As a \textit{type system} like this grows deeper and denser, more information is required to represent it. Similarly, any class in the system \textit{gains information} as it is specialized and \textit{loses information} as it is generalized. Thus, a class of type $T$ is \textit{also} of any supertype of $T$ because each possible supertype represents a subset of $T$'s information. For example, the class \textit{Car} is of type \textit{Car}, but it is also of the more general types \textit{Vehicle} and \textit{Machine}. \\

Naturally, one wonders about what might happen at the extremes of generalization and specialization. Three relevant question-answer pairs are given below: \\

\begin{itemize}
	\item \textit{Is there a type so general that it describes everything?} \\\\	
	Indeed, there is: it is the \textit{top type} $\top$, and it describes a special class known as a \textit{domain of discourse} or \textit{universe}. The names speak for themselves---the supertype of \textit{everything} describes the class of \textit{all things} under discussion, the \textit{universe} of relevant topics, the \textit{domain} of our discourse. $\top$ serves as a useful label in situations when type is unimportant or indeterminate.
	\item \textit{Is there a type so specific that it describes \dots\ nothing?} \\\\
	Yes. It is the \textit{bottom type} $\bot$, and it describes the \textit{empty class}, the class with zero members. It can be considered the type of an \textit{overspecified} class. That is, the class has been defined such that no object in the \textit{domain of discourse} complies. $\bot$ describes things that are \textit{out of scope} or \textit{nonexistent}. Thus, it is often used as the return type of functions that \textit{diverge} (i.e. do not terminate properly) or as the type of \textit{nothing} (e.g. the $5^{th}$ element in a $4$-element list is nothing).
	\item \textit{Are instances just classes that are specialized until they have one element?} \\\\
	Technically, no. Specialization is a class-to-class process: a class is specialized to a subclass. \textit{Instantiation}, the construction of a concrete \textit{instance} from an abstract description, is instead a class-to-object process. For example, in the above diagram, the class $Y_3$ is \textit{instantiated} in order to produce the object $b$ (or, by reversing the direction of the arrow, we may say that $b$ is an \textbf{instance of} $Y_3$). \\\\
	Because a type describes a class, it is also incorrect to say that $b$ defines its own subtype of \textit{Car}. Rather, $b$ is a \textit{token} of type \textit{Car}, a physical object that \textit{realizes} one of the many possible designs that fall under the class \textit{Car}. However, $Y_3$ could be specialized such that $b$ is the sole element of the resultant class ${b}$. A class with one member is called a \textit{singleton class}, and it has a \textit{unit type}, a type with one valid value. The type of ${b}$ would be a subtype of \textit{Car} that describes only $b$, but neither ${b}$ nor its type are the same thing as the \textit{instance} $b$. \\
\end{itemize}

\subsubsection{Ontologies}

% The difference between a taxonomy and an ontology is that the former has child nodes. Ontologies covers more than just is-a relations.

% Generalization is powerful because it allows you to know things about systems you haven't seen before. It gives intuitive understanding.

% This book takes a top-down approach (i.e. from a most general class of information down to all the different kinds of computational phenomena). First, we need to characterize the concept of information, which developed out of philosophy and physics.

% We gotta form a theory of information bottom-up by observing some natural systems (category systems, physical systems). Then, we discuss how systems and models are classified. Finally, we can apply the theory top-down in order to model the many systems involved in computation (language systems, logical/formal systems, algebraic systems, automatic systems, hardware systems, software systems, operating systems). All of these can be modeled in terms of information.

% Bottom-up is like starting from an arche, top-down is like starting from henosis

% Natural and constructed systems. Open-world, closed-world assumption.

% Computer science is a unifier of academic fields. It is hard to fully grasp because it is so general. We need to pull from many disciplines to characterize it effectively (language, logic, philosophy, math, physics, engineering).
% The theory of computer science was built many to one. Practical programming is one (information, bits) to many (types, software). Thus, we will discuss many different fields and will unify them under a theory whose arche is information (logical values).
% We must consider open-worlds before we discuss closed-worlds.

% We will discuss upper ontologies next. This will inform later discussions on domain ontologies for physics and computer science.

%----------------------------------------

\subsection{Classical Metaphysics}

% What is metaphysics? What is ontology? What is an ontology?
% What does metaphysics and ontology have to do with OOP?

Blah

% One could argue that philosophy has been done ever since we could think. But philosophy in Ancient Greece was different. Why? Geography. Democracy.
% Yes, other cultures had "philosophies" at this time, but the word in this case is referring to a \textit{worldview}. We are discussing philosophy in the sense of \textit{analysis}.

\subsubsection{\textit{Henosis} and \textit{Arche}}

Blah

% Motivation for metaphysics. The One. The Good. Henosis (unity). Arche (first principle).
% Physis (nature, growth) vs. nomos (human convention). The distinction was made maybe because the Greeks had contact with many different cultures interpreting the world in different ways.
% Thales. Explaining things without the supernatural. From many things to one substance (the arche is water). Cultural reasons for choosing water. A distinction between appearance and reality. You can live life according to appearances, but you can also learn for the sake of learning. And it turns out learning can be useful, too (olive story).
% Heraclitus. The arche is fire because it symbolizes change. Flow of time.
% Empedocles invents the classical elements.
% Rebooting Hesiod's Theogony. Instead of gods emerging from the Chaos, let's try to categorize everything naturalistically.

\subsubsection{Democritus' Atomism}

Blah

% Time is an instance of the Form of eternity. The Greeks did not think about time like us.
% Atoms come in different shapes that give different properties. The shapes also influence sensory perception (bitter/salty taste comes from sharp atoms).
% Materialism, mechanism, naturalism

\subsubsection{Plato's Theory of Forms}

% Plato hates Democritus. Wants to burn all of his books. Plato's philosophy is an objection to atomism. Believes that there must be something divine about nature due to its beauty.
% Plato is rationalist, Democritus is mechanist and materialist (they agree on epistemology though)
% Essence precedes existence. There is inherent meaning in reality.

Plato (c. 428/427 -- c. 348/347 BCE) was the first to expound a metaphysics of objects with his \textit{theory of Forms}, which proposed a solution to the \textit{problem of universals}. This problem is subtle, but important: how can one general property appear in many individual objects? For example, when we say that a floor and a bowl are \textit{smooth}, are we referring to some singular paradigm of \textit{smoothness} that is independent of each? For Plato, \textit{universals} or \textit{Forms}, such as smoothness, are indeed distinct from the \textit{particulars} that \textit{partake} of them, such as floors and bowls and other smooth objects. He posits that the \textit{essence} of an object, its \textit{sine qua non}, is the result of its partaking of these Forms. \\

\textit{Forms}, as conceptualized by Plato, are ideal, abstract entities that exist in the \textit{hyperouranos}, the realm beyond the \textit{heavens}. In the times of Homer and Hesiod, the Ancient Greeks understood a cosmology that closely mirrored that of the Hebrew Bible. They envisioned the Earth as a flat disc surrounded by endless ocean and the sky as a solid dome with gem-like stars embedded in it, much as it all appears from a naive perspective on the ground. Under the Earth was the \textit{underworld}, where the dead live, and beyond the dome were the \textit{heavens}, where the divine live. \\

% Elaborate on why spheres were chosen. The Universe was also a sphere.

Plato, however, was influenced by the philosophy of Pythagoras (c. 570 -- c. 495 BCE), who championed a spherical Earth for aesthetic reasons. Thus, he instead modeled the Earth as a "round body in the center of the heavens," and he imagined the Forms as part of a different, eternal world that transcends physical space and time. Whereas physical reality is the domain of perception and opinion, the realm of the Forms is, as he paints it in \textit{Phaedrus}, the domain of \textit{reason}: \\

\begin{displayquote}
	\textit{Of that place beyond the heavens none of our earthly poets has yet sung, and none shall sing worthily. But this is the manner of it, for assuredly we must be bold to speak what is true, above all when our discourse is upon truth. It is there that true Being dwells, without colour or shape, that cannot be touched; reason alone, the soul's pilot, can behold it, and all true knowledge is knowledge thereof. Now even as the mind of a god is nourished by reason and knowledge, so also is it with every soul that has a care to receive her proper food; wherefore when at last she has beheld Being she is well content, and contemplating truth she is nourished and prospers, until the heaven's revolution brings her back full circle. And while she is borne round she discerns justice, its very self, and likewise temperance, and knowledge, not the knowledge that is neighbor to Becoming and varies with the various objects to which we commonly ascribe being, but the veritable knowledge of Being that veritably is. And when she has contemplated likewise and feasted upon all else that has true being, she descends again within the heavens and comes back home. And having so come, her charioteer sets his steeds at their manger, and puts ambrosia before them and draught of nectar to drink withal.}
	\vspace{4mm}
\end{displayquote}

Plato makes a revolutionary statement here about the nature of scientific knowledge. He declares that one cannot know the physical world directly because it is, in fact, \textit{less real} than the \textit{hyperouranos}. As he puts it in \textit{Timaeus}, a Form is "what always is and never becomes," and a particular is "what becomes and never is." Further, physical reality is merely the "receptacle of all Becoming," a three-dimensional \textit{space} that affords spatial \textit{extension} and \textit{location} to all particulars. Unlike Forms, particulars are subject to \textit{change}, and thus cannot be sources of absolute knowledge. They are slippery, metaphysically undefinable, and explainable only in terms of the unchanging Forms of which they partake. \\

% Plato's criticism of Heraclitus' conception of time (wiki Heraclitus)

The above passage can be interpreted as an early argument for basing human knowledge in abstract mathematics rather than in subjective evaluations of concrete observations. For example, to \textit{know} what a triangle is, one cannot simply point to a triangular rock or even to a very carefully drawn diagram of a triangle. One must instead use \textit{abstract reasoning} to express the Form of a triangle by means of a \textit{formal definition}: a three-sided shape in a two-dimensional plane. This is the \textit{essence} of a triangle. In contrast, the triangular rock and diagram each partake of multiple Forms, and neither are perfectly triangular. We may still \textit{call} them triangles, but only those who know the Form of a triangle will understand what we mean when we do so. Therefore, it is knowledge of the unambiguous Form that gives insight into the complex and ever-changing particular, not the other way around. \\

Here, our object-property distinction blurs. Because while Forms may define universal properties, are they not also objects in an abstract sense? And while concrete particulars may be objects, are they not also fully characterized by a set of properties? In Plato's metaphysics, the categories of interest are really the \textit{concrete} and the \textit{abstract}, that which we \textit{perceive} and that which we \textit{conceive}. Plato further claims that humans come from the abstract realm, where they familiarize themselves with the Forms, and they are born into the concrete realm, where particulars are abundant and Forms are only latent memories. It is through philosophy then, the exploration of the rational mind, that we unearth our innate understanding. \\

To explain the link between the realms, Plato also postulates a \textit{demiurge}, a transcendent artisan who shapes an initially chaotic space into an ordered, material reality, using the Forms as models. This entity is benevolent and wishes to craft a world like the ideal one he sees, but the objects that he \textit{instantiates} are always imperfect copies. Like that of a master portraitist, his work is beautiful, but it is, in the end, a mere \textit{image} of his subject. \\

So it is for programmers. Before us lies the space of computer memory, uniform and featureless, a receptacle of Becoming. Our Forms are \textit{types}, our particulars are \textit{values}, and the behavior in the space is dictated by the \textit{code} that we execute. In \textit{object-oriented programming} (OOP), values can be of an \textit{object type} that is formally defined by a \textit{class}. \\

For example, my cat Mystic belongs to the class of entities known as Cat, the set of all possible cats. She is an \textit{object} known as a cat, and her catness is defined by the class Cat. In Plato's terms, she is a \textit{particular}, partaking of the Form \textit{Cat}. Just as the demiurge instantiates particulars from Forms, so the programmer instantiates objects from classes. Both bridge the gap between what is possible and what is. \\

That said, while Plato's metaphysics gives good insight into OOP and other abstract (or \textit{high-level}) programming concepts, it cannot explain concrete (or \textit{low-level}) ideas like \textit{data} that is represented with \textit{bits}. This is because Plato was not overly concerned with the \textit{substance} of objects, the stuff that makes them \textit{physical}. Thus, in our pursuit of a holistic understanding of \textit{information}, we must also consider \textit{substance theories}, philosophies that give metaphysical weight to \textit{matter}. \\

% Plato associates the classical elements with the Platonic solids. He argues that they are not atoms but corpuscles (sort of). Also, the elements are not the arche because they are material.
% Plato not very interested in matter. His arche is immaterial (triangles), not material (atoms).

\subsubsection{Aristotle's Hylomorphism}

Blah \\

% Aristotle is sort of a mixture of Democritus and Plato. Not entirely materialist, but not fully abstract. It's scientific, but also teleological. Democritus embraces physis, Plato rejects it, and Aristotle redefines it. Physis is defined by the four causes. Nature and art are both part of physis, but the latter requires an external force/motion. Thus, the natural world has a "soul."
% Aristotle did not have a singular arche. But he wrote about minima naturalia, the most fundamental of which were the matter + form combos of the classical elements. He did not believe in a void, reality was continuous (big effect on physics and mathematics).

% Substance != matter for Aristotle.
% Rejects atomism, says reality is continuous (corpuscles)
% Rejects an immaterial fundamental unit, believes substance is real
% Four formal causes
% Actuality and potentiality
% Objects are characterized spatially, events temporally.
% Mathematical objects and operators
% Plato is rationalist, Aristotle is empiricist

% "We (collectively, as humans) bestow globality on first principles inductively, promoting the said principles to the status of universal, often missing meaningful alternatives in the process." -- Alon Amit
% Put this after Euclid's section?

%----------------------------------------

\subsection{Theories and Models}

% Patterns

% Can we say anything objectively about the structure of the Universe? For example, can we say for certain that the Universe is continuous or discrete? Probably not. But we can build some very sophisticated theories and models that describe the Universe very well.

% Wiki - analogical models

% Axiomatic system -> theory -> model -> simulation

% System - a bunch of interrelated parts
% Theory - a way of looking at a system (syntactic, system-independent, but perhaps designed with the system in mind)
% Model - a way to represent a system (or the application of a theory to a system)

% Theoria vs. praxis (considering things outside of our control vs. considering things we can change)
% Pythagoras considered theories a way of contemplating math

% Scientific disciplines use math to describe parts of reality "in certain terms." Two scientific disciplines might describe the same thing from different vantage points. Theories as models. Scientific laws are not axioms. Models represent reality in some, but not all, ways. If they represented everything, they wouldn't be models. They would be the thing itself. In mathematical models of physics, statements represent physical phenomena.
% Theories that describe the same phenomena may differ based on what their basic abstract objects are (particles, waves, bits). These objects are reference points, helpful abstractions inspired by previous observation.
% Logic/math theories have axioms/postulates and theorems about formal phenomena and are interpretation-independent. Logic/math models are particular algebraic structures, abstract objects that satisfy a theory (e.g. a set S is a model of set theory, a group G is a model of group theory)
% Scientific theories have postulates and laws/principles about natural phenomena. Laws are classical and are expressed with formulas. They are not certain, but are rather extremely probable and backed by mountains of evidence. Principles are modern and are expressed in words. They are "guiding ideas" and do not require proof. Principles can be violated, but you better have good reason to suspect that's the case.
% Axioms vs postulates. Essentially synonymous, but the former is more common in math and the latter is more common in science. Mathematicians do not consider axioms verifiable. In science, postulates are about the physical world, and physics is verifiable, so postulates are verifiable. Postulates can be broken if contrary evidence is found.
% Models do not *exist*. They are mathematical abstractions that describe reality. Some phenomenon might be "well-described" by a theoretical model, but that does not make it a signal, field, etc.
% Table for model classifications: linear/non, static/dynamic, explicit/implicit, discrete/continuous, deterministic/probabilistic, deductive/inductive/floating
% Models may die out if they have major flaws. However, crude models are still sometimes useful due to their simplicity (classical mechanics).
% Theories are abstract models, models whose components are symbolic.

% A model represents a system, a simulation represents its operation over time.
% A computer program can be interpreted as a simulation of some aspect of the Universe and, at the time, as a little Universe in and of itself. The difference between this universe and \textit{the} Universe is that we theoretically have access to all of the information in the former. Whereas we are babies stumbling around in the dark of physical reality, we are gods of the universes we architect in digital memory.
% Mindset (blue box?). You can think of programming as just sitting at a desk and typing a bunch of obscure mumbo jumbo. Or you can think of it as the means to simulate anything that you want.

% Universe and Domain of Discourse
% Open and closed world assumptions
% Let's discuss how humanity has modeled physics over the years.

Blah

\subsubsection{Euclid's \textit{Elements}}

Although arithmetic has been practiced since prehistory, theoretical mathematics was founded by the Ancient Greeks. The Greeks, in contrast to earlier peoples, applied \textit{deductive reasoning} to the study of mathematics. While sophisticated arithmetic, geometry, and algebra was done before this period by the Babylonians and the Egyptians, this work was all based on \textit{evidence} collected from the physical world. The Greeks were the first to recognize the need for \textit{proof} of mathematical statements using \textit{logic}. \\

% Euclid used axiomatic and constructive methods
% Axioms come from Aristotle, formal definitions come from Plato

While various proofs were written by Greeks in the centuries before his birth, Euclid of Alexandria is nonetheless credited with formalizing the concept of proof due to his use of the \textit{axiomatic method} in his groundbreaking treatise, \textit{Elements}. This work was the first of its kind, serving as both a comprehensive compilation of Greek mathematical knowledge and as an introductory text suitable for students. Additionally, it was a product of a synthesis of Greek thought, founded on the epistemology and metaphysics of Plato and Aristotle. \\

\textit{Elements} is divided into thirteen books. Books I--IV and VI cover \textit{plane geometry}, the study of shape, size, and position in two-dimensional space. Books V and VII--X cover elementary \textit{number theory} (classically known as \textit{arithmetic}) in a geometric way, expressing numbers as lengths of line segments. Finally, Books XI--XIII cover \textit{solid geometry} by applying principles of plane geometry to three-dimensional space. Each book begins with a list of \textit{definitions} (if necessary) that assign names to abstract mathematical ideas that are expressed in unambiguous terms. These ideas are then used to prove \textit{propositions}, mathematical statements that warrant proof. \\

Euclid's proofs in \textit{Elements} can be described as \textit{axiomatic}. That is, they start from \textit{axioms}, statements that are taken as true without proof. Axioms solve a problem in logical argumentation known as \textit{infinite regress}, which often occurs, for example, when children ask questions. A curious child, seeking knowledge, might ask his parent why something is the way it is. That is, he requests a proof of some proposition $P_0$. The parent would respond with some proposition $P_1$ that supports the truth of $P_0$. The child, unsatisfied, would subsequently ask why $P_1$ is true. The parent would respond with the assertion that $P_2$ is true and implies $P_1$, and the child would similarly question whence $P_2$ arises. This dialogue could continue in this manner forever with the child endlessly searching for an anchor with which he can ground his understanding. However, patience waning, the parent must end this line of questioning by giving an unquestionable answer. Common examples include "Because that's just the way it is" or "Because I said so." \\

Some Greek philosophers considered infinite regress to be a serious epistemological issue. If we assume that all knowledge must be demonstrable by a proof or logical argument, we cannot know anything because our axioms cannot be satisfactorily proven. And yet, there are things that we claim to know. In his \textit{Posterior Analytics}, Aristotle challenges this notion that all knowledge must be provable, averring that \\

\begin{displayquote}
	\textit{All instruction given or received by way of argument proceeds from pre-existent knowledge.}
	\vspace{4mm}
\end{displayquote} 

Aristotle continues his text in support of this statement. He discusses the nature of \textit{premises} in his \textit{syllogistic logic} and how they differ from the \textit{conclusions} that they derive. Namely, he states that the premises of a knowledge-producing deductive argument or \textit{syllogism} must be \textit{true}, to ensure the truth of their conclusion, \textit{indemonstrable}, to ensure that they are independent of proof, and \textit{causal}, to ensure that their conclusion logically follows: \\

\begin{displayquote}
	\textit{Assuming then that my thesis as to the nature of scientific knowledge is correct, the premises of demonstrated knowledge must be true, primary, immediate, better known than and prior to the conclusion, which is further related to them as effect to cause. Unless these conditions are satisfied, the basic truths will not be 'appropriate' to the conclusion. Syllogism there may indeed be without these conditions, but such syllogism, not being productive of scientific knowledge, will not be demonstration. The premises must be true: for that which is non-existent cannot be known---we cannot know, e.g. that the diagonal of a square is commensurate with its side. The premises must be primary and indemonstrable; otherwise they will require demonstration in order to be known, since to have knowledge, if it be not accidental knowledge, of things which are demonstrable, means precisely to have a demonstration of them. The premises must be causes, since we posess scientific knowledge of a thing only when we know its cause; prior, in order to be causes; antecedently known, this antecedent knowledge being not our mere understanding of the meaning, but knowledge of the fact as well.}
	\vspace{4mm}
\end{displayquote}

Thus, Aristotle differentiates \textit{pre-existent knowledge} from \textit{scientific knowledge}. Whereas the latter is produced by means of a logical proof, the former is produced by one's \textit{intuition}. He goes on to argue that what we intuitively understand is indeed knowledge, in the same way that what we prove is knowledge. The difference, he argues, is that axioms are \textit{self-evident}. For example, the reflexive property of the natural numbers (i.e. for every natural $x$, $x=x$) is typically considered so obvious that proof is unnecessary. At the same time, the statement is so fundamental that it cannot really be proven. Despite their unprovability, Aristotle argues that axioms such as these constitute knowledge that is simply, unquestionably \textit{known}: \\

\begin{displayquote}
	\textit{Our own doctrine is that not all knowledge is demonstrative: on the contrary, knowledge of the immediate premises is independent of demonstration. (The necessity of this is obvious; for since we must know the prior premises from which the demonstration is drawn, and since the regress must end in immediate truths, those truths must be indemonstrable.) Such, then, is our doctrine, and in addition we maintain that besides scientific knowledge there is its originative source which enables us to recognize the definitions.}
	\vspace{4mm}
\end{displayquote}

Euclid apparently agreed with this epistemology because Book I of \textit{Elements}, in addition to its definitions and propositions, contains a number of axioms, which are labeled either as \textit{postulates} or \textit{common notions}. In antiquity, a postulate denoted an axiom of a particular science, whereas a common notion denoted an axiom common to all sciences. Today, we do not think of axioms in this way. Rather, we simply consider them starting points for proofs. Euclid's axioms are given below: \\

\begin{center}
	\textbf{Postulates}
	\begin{enumerate}
		\item Hi
	\end{enumerate}
	\textbf{Common Notions}
	\begin{enumerate}
		\item Hi
	\end{enumerate}
	
\end{center}

Euclid's proofs can also be described as \textit{constructive}. \\

% Aristotle's logic was two-valued. However, he did not fully accept the law of excluded middle.

% The Greeks used geometry to get around the precision errors of irrational numbers. That's why they used line segments to teach arithmetic.

% Elements is great and super important, but in some ways it held mathematics back. It was taught for over 2000 years, and many people assumed it could not be wrong. In fact, it is not considered formal by modern standards, and it is not a general theory of geometry. Critique of the 5th postulate. Non-Euclidean geometry, postulates of special relativity, Galois theory, modern mathematics does not treat axioms as "self-evident" but rather as starting points. Elements assumed many things to be generally true, which are not: numbers are one-dimensional, mathematical objects must be constructed, mathematics must describe the world as we observe it (2d/3d space, concrete objects), mathematics requires straightedge and compass construction, there is only one way to describe geometry. None of these are true, and modern mathematics began when people started questioning Elements.

% Aristotle's metaphysics influences science for 2000 years.

% Aristotle's ten categories
% Moving away from Aristotelian realism, Kant supposes that categories say something about how we conceptualize reality, not about reality itself.

%----------------------------------------

\subsection{Modeling Physical Reality}

% Religion, mythology, cosmogomy
% We're not going to deeply discuss religion or any theory that says more about the supernatural than the natural. However, we will briefly touch on some ancient models of physical reality that are intertwined with spirituality.
% Touch on religion up front, so we don't have to talk about it after this.
% Models are constructed within the "religion" of a society. This is not religion in the customary sense, but a spiritual, cultural bias, a set of guiding ideals. Heisenburg calls it the "spiritual pattern of a community." Religion in this sense does not have to be theistic.
% "Two kinds of truth"
% Religion helps us make sense of the chaos. A common theme is a belief in The One, an interconnection of seemingly disparate things.
% The progress of science was held back by war, disease, and also religion. But religion is not inherently destructive. Science and religion discuss two entirely separate things. The mistake was in trying to fit science into a religious model. Too many scientists spent time looking for rigid patterns in the Universe that meshed with their religion, and they just were not there. The Universe is not a divine puzzle that holds the meaning of the divine.
% In the same way that there are scientific things that religion has no bearing on, there are religious things that science has no bearing on (ethics). Science will not make religion obsolete. Humans need meaning, community, fulfillment, guidance, etc. Technology does not solve these things. Religion does (however, again, it doesn't require God).
% We see a trend from anthrocentric to not. We are biased toward our species, our intelligence, our planet, our culture, our language.

\subsubsection{Pre-scientific Thought}

% Atomic theory of Hindu philosophy
% Tao, Yin and Yang
% Indra's net
% Monad, Pythagoreanism
% Classical elements, aether
% Kosmos and Nous, Plotinus
% Pre-Socratic atomist philosophy
% And then came Socrates, Plato, and Aristotle
% Rome comes and goes, classical knowledge is lost to the West for hundreds of years, but is preserved in Byzantium/Constantinople (Nova Roma)
% Christianity dominates the Western world and academic work is done primarily in order to support Christian theology. Naturalism dies out.
% Some scientific work is done in the West during the Middle Ages, but it is isolated and there is little progress.
% Islamic Arabs seek out and acquire the knowledge of other civilizations, Classical Greek knowledge is acquired from Constantinople, Translation Movement, the Islamic Golden Age occurs, ends due to the destruction of Baghdad by Mongols
% Europe slowly recovers Ancient Greek knowledge, and academic work is done to harmonize Aristotelian thought with Christianity
% The Black Death happens and wipes out half of Europe, scientific work halts
% Constantinople falls to the Ottomans in 1453, Byzantine scholars flee to Europe, bringing their knowledge of algebra
% The invention of the printing press allows for the wide distribution of information
% A major religious schism occurs in Western Christianity (Protestant Reformation), prompting rebellion in science as well
% Copernicus publishes De Revolutionibus in 1543 on his deathbed
% Galileo champions heliocentric theory and is persecuted by the Church for it. "And yet it moves."
% The Scientific Method (Bacon, Decartes, Galileo, Newton) (wiki history of sm)

\subsubsection{Point Particles and Forces}

% Descartes publishes the idea of a "frame of reference" in 1637 (Fermat also discovers it).
% Descartes founds the rationalist, mechanistic philosophy that early models adhere to (1644). Newton publishes the Principia in 1687.
% Descartes develops the idea of a system.
% Basic assumptions based on observation: continuity, determinism, casuality, forces result from collision, absolute space and time
% Mechanics - study of the behavior of point/body/bodies subjected to forces
	% Statics - the net force is 0, the net work is 0, the body does not accelerate (at rest or constant velocity)
	% Dynamics - nonzero net force, nonzero work, the body accelerates
		% Kinematics - study of motion of a body without considering cause of motion (no masses, no forces, just motion)
		% Kinetics - take cause into account
% In Newton's era, physics was described in terms of bodies, things with mass (atoms, particles, objects). Scientists followed a mechanistic philosophy (physical reality is simply the motion and collision of matter, the universe is deterministic). Newton was a mechanist as well, but made an exception for gravity, which exhibited "action at a distance." He did not attempt to explain gravity, but simply argued that it matched the data.
% Newton discovered that objects would continue their motion if absent of friction, which is a pretty unintuitive idea, considering that there is always friction on Earth.
% Optics - Descartes says light is emitted by sources, Huygens develops a longitudinal wave theory of light transmitted though aether, but people don't believe it because it did not explain birefringence, Newton develops corpuscular theory of light and people accept that.

% Blue box for Hume's Problem of Induction and Kant's Trancendental Aesthetics.
% Hume states that induction is not infallable and thus is unjustified. Kant argues that while it is not infallable, it is still necessary and rational. Kant backs up the scientific method but argues that it does not constitute an objective truth. Science is not certain, but that's okay.
% Critique of Pure Reason - there are other ways of being rational besides deduction
% We can only have knowledge of what appears to us, not of things in themselves.
% Synthetic a priori knowledge (the knowledge of metaphysics) is what allows us to compute mathematics in our heads. Math is necessary and universal, and math can connect a concept to a new concept other than itself (ampliative).
% Religious implications of transcendental idealism
% Kant thinks S+T are just human projections. Synthetic a priori. They require no observation and are projected onto surroundings.
% Kant: nothing about the object's location describes the object itself.
% Walking in a thick grove of trees and finding an overlook into the mountains. Your sense of spatial scope changes. Your mind readjusts.

\subsubsection{Waves and Fields}

% In early 1800s, people realize Newton's theory does not explain diffraction, interference, and polarization (physical optics). Young and Fresnel develop a transverse wave theory of light. This is still a mechanistic theory, and, thus, it required an aether (as discussed by Aristotle). People tried to understand how an aether could exist if the Earth moves freely through it. Aether drag and wind theories emerge.
% Similar theories involving mysterious substances ("imponderable fluids") were formed to explain heat and electricity. Faraday proposes that reality is made up of "lines of force." Maxwell agrees and develops an electromagnetic field theory, but he assumes an aether is required.
% Michelson-Morley experiment discredits the aether. Lorentz and Poincare forms a theory with a motionless aether that does not interact with matter. In this case, the aether was more of a reference frame. Length contraction and "local time."

% "Old quantum theory"
% Einstein introduces special relativity, which does not require the concept of aether. Time dilation instead of local time. This theory could be explained in terms of an aether though. The vaccuum of space is more like an aether than a true vaccuum.

% Poincare LET quotes

% Macroscopic physics is now modeled with relativistic mechanics. It is usually described with Lagrangian/Hamiltonian mechanics/field theories.

\subsubsection{Quantum States and Operators}

% Modern physics (macro and micro) is described with the Standard Model (quantum mechanics + QFT). This model is imperfect.
% Relativistic, quantum, and quantum field models are applied in different situations depending on the scale of space and time involved. Relativistic mechanics encompasses classical mechanics. It's just more complicated, so classical is still used by humans in situations where relativistic effects are negligible. Scientific computing has no reason to use classical though.
% Energy is transfered due to colliding particles or radiation. The former is modeled by particle theories, the latter by field theories.
% Mass-energy equivalence. Stationary objects have energy because mass is a manifestation of energy. Particles are not necessarily pieces of mass. They are more generally pieces of energy (quanta).
% "The principle of the conservation of mass [...] proved inadequate in the face of the special theory of relativity. It was therefore merged with the energy conservation principle—just as, about 60 years before, the principle of the conservation of mechanical energy had been combined with the principle of the conservation of heat [thermal energy]. We might say that the principle of the conservation of energy, having previously swallowed up that of the conservation of heat, now proceeded to swallow that of the conservation of mass—and holds the field alone. - Einstein
% Time dilation, causality
% Mass and energy (momentum) describe the same thing in different ways. Physics was originally skewed toward the mass view because it is more intuitive and concrete for humans.
% In general, outdated models tend to be anthropocentric in scale for space, time, and semantic meaning (theology). However, less general theories are not "broken." They are still useful in certain scenarios (domains of applicability).
% At small scales, things are unpredictable. They don't really follow laws. Things happen with certain probability. Classical laws are just averages of tons these probabilities from the same distribution, and thus, at human scale, objects behave basically the same way every time. The probability average is extremely close for each test.
% "You can't step into the same river twice." Likewise, classical physics does not behave in exactly the same way twice. We do our best to show general trends, but the Universe is not certain.
% In some sense, quantum mechanics supports object/event metaphysics. Fermions and bosons. Quantum fields of mass and force. Spacetime. Mass-energy. Identical particles.
% QM introduces a conflict over scientific realism. Should scientific theories strive to literally describe reality?
% Because QM is probabilistic, this implies that physics is nondeterministic. According to QM, we will never know the true nature of reality with certainty. We cannot know everything about every elementary particle.
% Problems with quantum entanglement. With the Copenhagen interpretation, we maintain that models are NOT reality. Just because the math asserts that some real physical quantity has some value, this is not certain without actually performing a measurement. Theories cannot predict values with absolute certainty.
% EPR paradox. Bell's theorem. QM seems to say things that are impossible. Throwback to Galileo: "And yet it works."
% Copenhagen interpretation claims that QM is a way of acquiring knowledge, but it does not attempt to describe objects in the way that classical mechanics does. It asserts that founding a model on entities or distinctions (such as particles, waves, and bodies) that are intuitive but ultimately unknowable, leads to incorrect predictions somewhere. Instead, it tries to make claims only about things that can be empirically observed.
% Conjugate variables (mass and momentum, uncertainty principle). Derivatives of action. Fundamental interactions. 
% Beyond the standard model: supersymmetry, integrating gravity: string theory, gravitons

% But there is an alternative way of looking at physics that is quite different from the models presented so far...

%----------------------------------------

\subsection{Information Theory}

% A pattern as defined in terms of information (Alon Amit). We study patterns in math and use them to model what we observe in physics. All patterns in the physical world can be described by information.

% How would you describe a beautiful gown with information?

% Bits and bytes and other units of information.
% The base-2 metric system (kibi,mibi,gibi,...)
% The scale of objects - an essay, a book, an encyclopedia, a 1080p photo, a 4K photo, a 1 minute 1080p video, a 1 minute 4K video, a 1080p movie, a 4K movie, an operating system (Linux vs Windows size), programs (little clients to massive environments like modeling softwares), etc.

% If physical phenomena are interpreted as information, this gives formal meaning to the philosophical idea that "we know very little." Much more physical information goes unnoticed than noticed.

% Information is similar to energy transfer. It was compared to theories of heat transfer done during the industrial revolution. When you do work, you give some of your energy to something else. The energy may also be interpreted as changing type (kinetic, chemical, etc.). However, this is also just an abstraction. Energy is energy. There are no types really.
% For example, when you throw a baseball, you transfer chemical energy in your body to kinetic energy in your arm and then you transfer that kinectic energy to the ball. During energy transfer, some of the energy becomes heat. But really, it's all the same energy. You transfer energy to the ball, and most of the energy manifests in motion, but some of it manifests in heat. Efficiency, then, is the ratio of how much of the energy involved is doing what you want it to.

% Entropy - transformational content of energy
% Boltzmann entropy, thermodynamic entropy, von Neumann entropy

% Mathematical theory of communication, Shannon-Weaver model
% What does it mean to be a good communicator?

%----------------------------------------

\subsection{Signals and Systems}
% Signals are messages composed of time-amplitude data

Data are the building blocks of \textit{signals}, which are mathematical models of quantities that change over time. Signals are carriers of \textit{information} about these quantities. A piece of information, known as a \textit{message}, is encoded as data of a certain \textit{medium}, either physical or numerical. In this sense, one can consider information a meaningful pattern that is embedded in a phenomenon. Philosophically, information is the \textit{essence} of a \textit{substance}, a set of properties that defines what a thing is. Mathematically, signals are functions $f$ that map a set of times $t$ to a set of amplitudes $A$, and they are classified according to the properties of these sets. \\

Sample rate is a property that is related to time. \textit{Continuous} signals are functions of continuous time. That is, $t$ is either the real number line $\mathbb{R}$ or an interval of it. In this case, $t$ is also called a \textit{linear continuum}. This definition of continuity is different from those given in real analysis, which describe functions for which sufficiently small changes in input result in arbitrarily small changes in output. For this reason, these signals are also called \textit{continuous-time} in order to avoid any confusion with the continuous functions that are described in mathematics. If a signal is not continuous, it is \textit{discrete} or discrete-time. \\

Continuous signals that have an infinite number of possible amplitude values (i.e. those that are modeled by $\mathbb{R}$ or $\mathbb{C}$) are called \textit{analog} signals because an infinite-valued output can be viewed as \textit{analogous} to the seemingly infinite complexity of real-world phenomena. At any instant in its time domain, an analog signal can \textit{exactly} model a time-varying, real-world quantity, such as voltage, pressure, distance, mass, color, or anything else that is quantifiable. All continuous signals are analog, but not all analog signals are continuous. For example, the analog data of a continuous signal can be sampled at a finite rate in order to construct a discrete, analog signal. \\

% ------

In general, signals can be viewed as messages encoded in \textit{data} whose decoded content is \textit{information}. Like traditional written messages, ... Any signal can be expressed as an ordered sequence of data, each of which can be expressed as a tuple $(t,A(t))$. Because there are an infinite number of "instants" in any given interval of continuous time, the continuous signal shown above can be expressed as an \textit{infinite}, ordered sequence of analog data. \\

% Vibrations, oscillations, music

Consider an arbitrary continuous signal:

% Make the signal weave on either side of the axis? Could be cool, but maybe distracting.

\begin{center}
	%\resizebox{\textwidth}{!}{%
	\begin{tikzpicture}
	\begin{axis}[
	width=\textwidth,
	height=5cm,
	tick style={draw=none},
	axis line style={draw=none},
	xticklabels={},
	ytick={0},
	y label style={rotate=270},
	ylabel={$A$},
	domain=-1:13,
	ymin=-5,
	ymax=5,
	]
	\addplot [mark=none, black] {0};
	\addplot [mark=none,
	samples=100,
	textbook-blue-dark,
	ultra thick,
	restrict x to domain=0:12] {2*sin(deg(x))+3*sin(deg(3*x))+2*sin(deg(7*x))};
	\end{axis}
	
	% Update the arrowheads, pls
	\fill (11.15,1.71) -- ++(-0.16,0.1) -- ++(0,-0.2);
	\draw (11.7,1.71) node {$t$};
	
	\draw (-0.22,2.2) -- ++(0,1.5);
	\draw (-0.22,1.2) -- ++(0,-1.5);
	\fill (-0.22,3.7) -- ++(0.1,-0.16) -- ++(-0.2,0);
	\fill (-0.22,-0.3) -- ++(0.1,0.16) -- ++(-0.2,0);
	\end{tikzpicture}
	%}
\end{center}

Analog data are special because they exist in continuous space and thus theoretically have an infinite degree of precision. If we assume a classical model of physics in which time flows continuously, we can model physical quantities with real numbers by writing an equation $A=f(t)$ where $t,A\in\mathbb{R}$. In this sense, a complete set of analog data over a given time interval (i.e. an analog signal) can perfectly describe a physical quantity over the same interval. However, handling analog data can be very difficult. \\

Imagine that you throw a ball in the air and somehow exactly measure its height above the ground at every instant before it lands. If you were to organize these height data with respect to time, you would construct an analog signal that perfectly describes the height of the ball at any point during the throw. In practice, however, no measuring instrument has the required speed or precision necessary to do this. Additionally, you would need an infinite amount of memory to store continuous-time data that is collected at every instant. \\\\

% Need digital data

\begin{tcolorbox}[breakable, enhanced, colback=textbook-blue, sharp corners]
	\vspace{3mm}
	\begin{center}
		\textbf{The Rationale for Binary Numbers}
	\end{center}
	% Implementation details: signals, logic levels, noise, power, reliability
	% Close relationship with classical, two-valued logic
	
	Binary encoding of data is not a requirement for computation. Rather, it is a concession that is made toward digital electronics. Computers have been designed for other bases, and, in some cases, a non-binary encoding is actually preferable. As long as the data can be encoded with a finite number of symbols, it can be handled by a \textit{digital} computer (i.e. a computer that manipulates digits). A digital computer receives and operates on a \textit{discrete} signal of data, a value or set of values that updates at regular time intervals (e.g. every second, every minute, etc.). \\
	
	One could argue that \textit{analog} computers process data that are encoded with an infinite number of symbols. However, this would require a loose definition of "symbol" because analog data cannot be perfectly represented by digits. Rather, analog data is represented in a real, physical medium, like position, pressure, or voltage. An analog computer receives and operates on this \textit{continuous} signal of data that updates instantaneously with real-world time.  \\
	
	Consider the  \hrefcolor[blue]{https://www.youtube.com/watch?v=s1i-dnAH9Y4}{\underline{Mark 1A Fire Control Computer}}, a device, more appropriately termed a "calculator" by modern definitions, that was installed on United States battleships in World War II. This machine used mechanical parts, such as gears, cams, followers, differentials, and integrators, to perform the arithmetic and calculus necessary to hit an enemy ship with a projectile, even if both ships were moving in different directions. It also used smooth components that gradually translated rotational motion into linear motion along a continuum. \\
	
	Despite being labeled with decimal (base-10) integers, this tool actually received \textit{continuous}, \textit{analog} quantities, such as physical position, as input. If each of its mechanical positions were considered a different "symbol," this system would theoretically be able to represent uniquely every real number to a infinite degree of precision and, thus, compute continuous functions without error. Such analog computation is performed also in slide rules, as well as in early electronic computers, which used voltage to represent numbers instead of position. \\
	
	\begin{center}
		\begin{tikzpicture}
		% Ruler
		\draw [thick] (0,0) rectangle (10,1);
		\draw (0,0.5) -- ++(7.2741,0);
		\fill (7.2741,0.5) -- ++(-0.16,0.1) -- ++(0,-0.2);
		\foreach \i in {0,...,10} {
			\draw (\i,1.6) node {$\i$};
		}
		
		% Slider
		\draw [thick] (7.2741,-0.2) rectangle (8.2741,1.2);
		\draw (7.7741,-0.2) -- (7.7741,1.2);
		\fill (7.7741,0) -- ++(-0.1,-0.2) -- ++(0.2,0);
		\fill (7.7741,1) -- ++(-0.1,0.2) -- ++(0.2,0);
		\draw (7.7741,-0.7) node {$7.7741$};
		\end{tikzpicture}
	\end{center}
	
	% Analog computers have many problems: noise, reproducability
	
	Any digital encoding of numbers is possible in computing, and the choice does not affect what a computer theoretically can or cannot do. Rather, it has practical consequences. Decimal encoding was implemented in a number of early electronic and electro-mechanical computers, such as the Harvard Mark I and the ENIAC. 
	
	% Ternary electronic computers and their issues
	
	% Binary computers and boolean algebra (logic)
	
	% Beyond binary - fuzzy logic (briefly, discuss more deeply later)
	\vspace{3mm}
\end{tcolorbox}
\vspace{2\baselineskip}

% Encoding doesn't have to use numerals. It can use letters, pictures, whatever. You can encode anything anywhere.

% Information signals
% Information systems

Regardless of which encoding is used, data is data.

% All information is encoded and decoded (digital and sensory)

% So, we can represent physical things we experience with digital data. If we collect enough data on the properties of these things, we can simulate reality. We can also simulate abstract mathematical phenomena and other abstract things like language and logic. Much of software design is the simulation of mathematical objects.

% Structured manipulation = algorithm
% The Shannon-Weaver Model of Communication
% Sensory, conversation, and machine examples

% We need to communicate an algorithm in a language
% What kind of language?
% First, let's talk about language in general

% Serialization or marshalling - translating an object (abstract or concrete) into a format that can be stored, communicated, and deserialized into an object again (terms are not quite synonymous in Java)

Data are unorganized facts or statistical values that may reveal something of interest. They are collected by \textit{sampling} the properties of \textit{phenomena} from the observable Universe. Once collected, they can be cleaned, organized, and then analyzed in the hope of gaining a better understanding of reality. When a set of structured data is put into context, it can be interpreted as \textit{information} that explains what has been observed. This information can then be \textit{encoded} as an ordered sequence of data known as a \textit{message}. Messages are \textit{transmitted} and \textit{received} both by Nature and by Man. In the latter case, this process can be viewed as the \textit{communication} of abstract, human thought. \\

Data can describe a variety of things, such as quantities, qualities, and relations. In computer science, data are \textit{quantitative}. That is, they describe \textit{quantities}, which are properties that can be modeled with \textit{numbers}. A quantity can be further classified as either a \textit{multitude} or a \textit{magnitude}. The former is a \textit{discrete} quantity that is \textit{counted}. One would ask "how many" elements belong to a multitude. In contrast, a magnitude is a \textit{continuous} quantity that is \textit{measured}. It is the size of something that is modeled as a \textit{continuum} (e.g. by fields such as $\mathbb{R}$ or $\mathbb{C}$), and one would accordingly inquire as to "how much" of it there is. \\

A set of data collected at a finite rate (e.g. by a \textit{measuring instrument}) is technically a  multitude of samples, even if the quantity being measured is a magnitude. For example, if a ball were thrown in the air, one could ask "how much" height it has at regular intervals of time. Once the data collection ends, one can also ask "how many" samples were acquired. Furthermore, one can ask "how often" samples were taken. This is an evaluation of the \textit{sampling rate}, and it is really just an alternative way of asking with "how much" speed samples were taken. Rates and ratios are still quantities. The existence of the rational numbers $\mathbb{Q}$ are evidence enough of this. \\

Data can be represented by various \textit{media}, both symbolic and physical. In the age of modern computing, data are often assumed to be \textit{digital} in nature, expressed in the symbolic medium of numerical \textit{digits}. This is the case for data processed by modern, electronic computers, which charge capacitors in order to generate voltages that represent \textit{logic levels}. A \textit{transistor} acts as a gate to each capacitor, staying closed to retain \textit{electric charge} during storage and opening to measure or modify voltage on reads and writes respectively. \\

The ratio of a capacitor's charge to its capacitance determines its voltage (i.e. $V=\frac{q}{C}$). This voltage is then mapped to a logic level, which corresponds to a range of voltage values. These logic levels represent digits. In modern computer memory, data are represented as binary digits or \textit{bits} (i.e. the digits 0 and 1), each of which represents a binary number. In this case, there are two logic levels, \textit{low} and \textit{high}. \\

\begin{table}[H]
	\centering
	\caption{Voltage Ranges for CMOS Logic Levels (V)}
	\label{tab:logiclevels}
	\begin{tabular}{|c|c|c|}
		\vtabularspace{3}
		\hline
		Signals & Low (0) & High (1) \\
		\hline
		Input & 0.00--1.50 & 3.50--5.00 \\
		Output & 0.00--0.05 & 4.95--5.00 \\
		\hline
		\vtabularspace{3}
	\end{tabular}
\end{table} 

However, quantitative data need not be expressed with symbolic digits. They can instead be defined in terms of an \textit{analogous} physical quantity. For example, the unit of measurement known as the \textit{millimeter of mercury} (mmHg) defines pressure in terms of the height of a column of mercury in a manometer. It is used to quantify blood pressure, and it does so by considering height an \textit{analog} or \textit{model} of pressure measuring that instead. Using this method, data about one phenomenon can be encoded as data of a different phenomenon, provided that the quantities involved are somehow related to each other. In this case, a height of 1 mm is related to a pressure of 133.322 Pa above atmospheric pressure. \\

% Computation with analog data means you are using actual, original, non-replicable quantites. Computation with digital data means you are using simulated quantities that are replicas or model of what you actually care about. We model EVERYTHING with numbers: physics, language, logic, images, video, music.

% Pros and cons of digital and analog computation

% One can think of digital encoding as a special case of analog encoding, in which a quantity is considered analogous to a sequence of digits that represents a \textit{number}. Because numbers are abstract and data must be concrete, numbers must be associated with symbols known as \textit{numerals} if they are to be used in computation.

%----------------------------------------

\subsection{Semiotics, Language, and Code}

% Types of sentences - declarative and imperative
% An algorithm as an argument

% Computation requires communication
% Model of communication
% All information is encoded and decoded (digital and sensory)

However, the term "universal language" is also often used in other contexts to describe a hypothetical language that all of the world speaks and understands. This is an old idea, and it is addressed in a number of myths and religious texts. In Genesis, the myth of the Tower of Babel explains the diversity of language as a result of God thwarting the Babylonians' plan to build a tower that would extend into the heavens. As punishment for their blasphemy, God "confused their tongues" and dispersed them across the world, shattering their once universal language. Similar myths are told by cultures across the world, such as the Native American Tohono O'odham people, who tell tales of Montezuma attempting to do the same, attracting the ire of the Great Spirit. \\

Alas, no such language exists and it is unlikely that one has ever existed. Rather, human language is generally believed to have evolved independently around the world from prelinguistic systems of communication such as gesturing, touching, and vocalization about 100,000 years ago. Similarly, written language evolved from proto-writing, the repeated use of symbols to identify objects or events. \\

Writing is distinct from speaking in that it is a reliable method of storing and transmitting information. Before the invention of written language, important pieces of history and literature were preserved through \textit{oral tradition}, passing from one generation to the next through repetition and memorization. However, oral tradition is prone to data loss and unintended changes. Like messages in a game of telephone spanning centuries, stories were at risk of losing old details and gaining extraneous ones. Additionally, if a society fell apart, their stories could be lost forever. Writing is a tool that mitigates these issues by \textit{encoding} speech into a symbolic code that can be inscribed onto durable media, such as clay tablets or stone, or onto more delicate, portable media such as parchment or paper. \\

% Quipu, number systems

\begin{tcolorbox}[breakable, enhanced, colback=textbook-blue, sharp corners]
	\vspace{3mm}
	\begin{center}
		\textbf{Symbology in Natural Languages}
	\end{center}
	% Symbol - syntactic, general term
	% Glyphs - syntactic, a graphical representation
	% Characters - semantic, associated with meaning (in computing, it refers to an "encoded character" that is represented by a glyph), a unit of information (which is interpreted by means of a type)
	% Graphemes - the smallest unit in a writing system, a written symbol that represents a sound or phoneme (a logogram, a syllabic character, a letter)
	% Writing systems - ideographic/pictographic systems (Emoji), logographic system, syllabaries, segmental scripts, alphasyllabaries
	% Non-linear alphabets, flag semaphore, animal communication
	% Character encodings
	% ASCII
	% Westernism - Scientific Revolution, alphabet, character, terms and jargon in English (covered in greater detail later)
	% Unicode, code points for abstract characters, UTF-8
	
	A \textit{symbol} is a syntactic mark that is understood to represent a semantic idea. For example, the numeral "$2$" is a symbol for the abstract concept of the number known as "two," the quantity of items in a pair. Symbology is universal for mankind, as it is an exercise of our capacity for abstract thought. Humans can leverage the pattern-recognizing power of the brain to associate symbols or sequences of symbols (also known as \textit{strings}) with ideas. Thus, we can proliferate information textually by means of a \textit{writing system}. \\
	
	Languages are diverse, and so are their writing systems.
	\vspace{3mm}
\end{tcolorbox}
\vspace{2\baselineskip}

% Constructed languages, a subset of which are formal
% Trade language, lingua franca, Latin
% Leibniz - Alphabet of human thought, "best of all possible worlds," panglossianism

An argument of any kind must be expressed in a \textit{language}. In the case of an informal argument, such as a debate, a \textit{natural language} is typically used (i.e. a language, spoken or written, that evolved \textit{naturally} as an aspect of human culture). This is done in the name of \textit{universality}. For example, a political debate that is held and perhaps broadcast in Poland would likely be conducted in Polish because Polish politics are primarily of interest to Polish people. Many Poles speak English in addition to Polish, but there is no language more widely spoken in Poland than Polish, with a whopping 97\% of the country declaring it as their first language. For media directed toward the population of Poland, there is no language that will better ensure an effective dissemination of ideas. It is a nearly \textit{universal} language in this context, a language understood by all intended recipients. \\

Formal proofs, on the other hand, require a \textit{formal language} for clarity and precision. In practice, formal proofs are rarely written or read by logicians or mathematicians. Typically, mathematical proofs are written using a combination of formal and natural language. However, computer programs, written in \textit{programming languages}, are considered formal. \\

A formal language is constructed from an \textit{alphabet} $\Sigma$, which is a set of symbols. The set of all possible finite-length \textit{words} or \textit{formulas} that can be built from this alphabet is denoted $\Sigma^*$. A formal language $L$ over an alphabet $\Sigma$ is then defined as a subset of $\Sigma^*$. Thus, a language is a set of purely syntactic objects. It may conform to a \textit{grammar} that specifies how words can be produced or arranged in a \textit{sentence}, but a language is never inherently meaningful. The semantics of a language is always interpreted separately from the syntax.

% ~~~ Numeral Systems ~~~

% and these numbers can be \textit{encoded} with either a single \textit{symbol} or a sequence of symbols called a \textit{string}. For example, the number eight is typically encoded with the symbol 8, which is a \textit{numeral} or \textit{digit}. The number could just as easily be represented by the symbol \varstigma, but this would be non-standard. Numbers are usually encoded according to an established \textit{numeral system} of some \textit{base} or \textit{radix} that specifies how many unique symbols are used for representation. Numerical data intended to be read by humans are typically encoded in a base-10 system, one of ten digits. \\

% This design choice is ancient and natural. It is based on the ten digits of our hands, which have aided us in counting for hundreds of millennia. In fact, fingers are the oldest counting tools used by man, and we still use them today for this purpose. Other anatomically inspired bases have been used in early numeral systems, such as base-20, which counts the fingers and the toes, or base-12, which uses a thumb on one hand to count the bones of the other four fingers. However, today, the base-10 \textit{Hindu-Arabic numeral system} is universal in mathematics. \\

% The base-60 system of Babylonian numerals is also notable. Its base is thought to have been chosen for its many factors, as this simplifies the process of division. Under this interpretation, this system is \textit{constructed} rather than natural. Instead of mirroring something concrete like physical extremities, this system chooses an abstract base in order to make a procedural optimization. The reasoning for base-2 encoding in computing is similar. It sacrifices human familiarity for performance and reliability. \\

%--------------------------------------------------------------------------------
%    SECTION: REASONING AND LOGIC
%--------------------------------------------------------------------------------

\toclineskip
\section{Reasoning and Logic}

% "Of course, the objects of a deductive system are normally thought of as formulas, the arrows are thought of as proofs or deductions, and operations on arrows are thought of as rules of inference."

% Laws of thought
% Three reliable ways of acquiring knowledge: deduction, induction, and communication.
% Hume's problem of induction
% Deductive: truisms via consequence, Inductive: facts via evidence

%----------------------------------------

\subsection{The Syntax and Semantics of Argument}

\textit{Reasoning} is a cognitive act performed by rational beings. It involves the absorption and synthesis of presently available information for the purpose of elucidating knowledge. It may require the ingenuity of thought, but in some cases the rote application of simple rules is sufficient. Reasoning could also be described as the providing of good \textit{reasons} to explain or justify things. However, as you might expect, there is much debate over whether or not any particular reason is "good." Reasoning is broad, and it comes in different flavors, but ultimately it is the pursuit of \textit{truth}. \\

Reasoning may involve the application of \textit{logic}. This \textit{logical reasoning} is the kind of reasoning that can be expressed in the form of an \textit{argument}. However, this argument does not need to be "correct." Such reasoning simply must be explainable in \textit{steps}, whether that be through informal speech or formal writing. \\

In logic, an argument has two parts: a set of two or more \textit{premises} and a \textit{conclusion}. If the conclusion \textit{necessarily} follows from the premises, the argument is called \textit{valid}. To know whether or not an argument is valid, one must produce a step-by-step \textit{deduction} of the conclusion from its premises. \\

A deduction is constructed within a \textit{deductive system}, which also has two parts: a set of two or more \textit{premises} and a set of \textit{rules of inference}. A rule of inference can be thought of as a kind of logical function. It takes statements (such as premises) as input and outputs either an intermediate result or the conclusion. A deduction, then, has \textit{three} parts: a set of two or more \textit{premises}, a \textit{conclusion}, and a set of \textit{intermediates}. A single-step deduction is represented below, along with a suitable rule of inference labeled $MP$. Note that the set of intermediates is empty in the case of deductions with only one step. \\[2mm]

% Modus Ponens diagram
\begin{center}
	\begin{tikzpicture}[scale=0.2]
	\tikzstyle{every node}+=[inner sep=0pt]
	
	\draw [black] (0,0) circle (3);
	\draw (0,0) node {$P$};
	
	\draw [black] (15,0) circle (3);
	\draw (15,0) node {$Q$};
	
	\draw [black] (3,0) -- (12,0);
	\fill [black] (11.2,0.5) -- (11.2,-0.5) -- (12,0);
	
	\draw [black] (0,-10) circle (3);
	\draw (0,-10) node {$P$};
	
	\draw [black, thick] (-9,-17) -- (18,-17);
	
	\draw [black] (0,-24) circle (3);
	\draw (0,-24) node {$Q$};
	
	\draw (-7,0) node {$1^\textit{st}$};
	\draw (-7,-10) node {$2^\textit{nd}$};
	\draw (-7,-24) node {$\textnormal{Con.}$};
	
	%----------------------------------------
	
	\draw [very thick] (26,3) rectangle (48,-27);
	\fill [textbook-blue] (26.1,2.9) rectangle (47.9,-26.9);
	
	\draw (37,-6.5) node {\fontsize{50}{40} $\textnormal{M}$};
	\draw (37,-18.5) node {\fontsize{50}{40} $\textnormal{P}$};
	
	%----------------------------------------
	
	\draw [black, thick] (22,3.07) -- (22,-13.07);
	\draw [black, thick] (22,3) -- (21,3);
	\draw [black, thick] (22,-13) -- (21,-13);
	\draw [black, thick] (22,-5) -- (26,-5);
	\fill [black] (25.2,-4.5) -- (25.2,-5.5) -- (26,-5);
	
	\draw [black, thick] (22,-20.93) -- (22,-27.07);
	\draw [black, thick] (22,-21) -- (21,-21);
	\draw [black, thick] (22,-27) -- (21,-27);
	\draw [black, thick] (22,-24) -- (26,-24);
	\fill [black] (22,-24) -- (22.8,-23.5) -- (22.8,-24.5);
	\end{tikzpicture}
\end{center}
\vspace{5mm}

This is the quintessential rule of inference used in logical reasoning, expressed here in symbols. It is known as \textit{modus ponens} (Latin for "a method that affirms"). The first premise $P\rightarrow Q$ is called a \textit{conditional statement} where $\rightarrow$ is the \textit{implies} operator. It states that if the \textit{antecedent} $P$ were true, it would \textit{imply} that the \textit{consequent} $Q$ would also be true. The second premise $P$ simply states that the statement $P$ is indeed true. Thus, by means of \textit{modus ponens}, we can \textit{infer} from the premises $P\rightarrow Q$ and $P$ that the conclusion $Q$ must be true. \\

Inference is also called \textit{entailment}. Though these terms describe the same logical concept, the latter more aptly describes the relationship between statements that are connected in a deduction. To infer is to come to a conclusion that is not \textit{explicitly} stated in the premises. To entail is to require that something follows. If $P$ \textit{entails} $Q$, this suggests that $Q$ is a \textit{logical consequence} of $P$, which we can express using a \textit{turnstile}: $P\vdash Q$. Inference is an action that humans perform. Entailment is a relationship between logical rules. \\

When the inferences made mirror the entailment, you've got yourself a \textit{valid} argument. However, this does not necessarily mean that your argument is \textit{sound}. \textit{Validity} is a property of arguments with a conclusion that can be \textit{deduced} from its premises according to rules of inference. However, validity says nothing about whether or not the premises have a \textit{meaning} that accurately describes reality. \textit{Soundness}, on the other hand, is a property of valid arguments whose premises are \textit{known truths} or \textit{axioms}. The deduction of a sound argument is called a \textit{proof}. \\

Validity and soundness complicate entailment. Now, an argument can be evaluated either in terms of its \textit{syntax} or in terms of its \textit{semantics}. Syntax refers to the arrangement of symbolic \textit{words} (or, more generally, \textit{tokens}) in a body of text. Semantics, on the other hand, refers to the \textit{meaning} that can be \textit{interpreted} from the text. Both of these concepts are integral to human language and indeed to the communication of information in general. \\

Consider the following valid deduction: \\

\begin{center}
	\begin{tabular*}{0.55\textwidth}{@{\extracolsep{\fill} } ll}
		\textnormal{"If it is raining, it is Tuesday."} & $P\rightarrow Q$ \\
		\textnormal{"It is raining."} & $P$ \\
		\hline
		\textnormal{"Therefore, it is Tuesday."} & $\therefore Q$ \\
	\end{tabular*}
\end{center}
\vspace{4mm}

This conclusion is a \textit{syntactic consequence} of the premises ($P\vdash Q$). That is, when the argument is evaluated simply as a collection of symbolic strings that are manipulated according to rules, $P$ entails $Q$. \\

Syntactic consequence is sometimes expressed with the following statement: "If the premises are true, the conclusion must also be true." However, while this statement does hold for all valid arguments, text evaluated syntactically does not have any notion of truth. It is simply a sequence of tokens. Thus, it may be better to think of valid arguments as "games that are played according to the rules." For example, chess has no inherent meaning associated with it, but it still has rules, and valid games of chess are those that comply with those rules. \\

Now, it might be Tuesday and raining as you are reading this. However, semantically, it is not required to be Tuesday if it is raining. The first premise given above is, in reality, false. However, if we change it to something that we consider true, we can come to a conclusion that is both valid and sound: \\

\begin{center}
	\begin{tabular*}{0.55\textwidth}{@{\extracolsep{\fill} } ll}
		\textnormal{"If it is raining, it is wet outside."} & $P\rightarrow Q$ \\
		\textnormal{"It is raining."} & $P$ \\
		\hline
		\textnormal{"Therefore, it is wet outside."} & $\therefore Q$ \\
	\end{tabular*}
\end{center}
\vspace{4mm}

In this case, the conclusion is a \textit{semantic consequence} of the premises ($P\vDash Q$). In addition to the argument being syntactically logical, it is also semantically true "in all universes." That is, there is no possible scenario where it is raining, and it is not wet outside. This syntactic-semantic difference will come up repeatedly throughout this guide. \\

%----------------------------------------

\subsection{The Structure of Proof}

Of course, arguments can have more than one step. An argument can involve many inferences, some of which may use \textit{intermediate conclusions} as premises for further conclusions. Multiple threads of reasoning can extend from the premises, weaving through intermediate conclusions and intertwining via inference until they all meet at a \textit{final conclusion}. This is the \textit{tree structure} that is inherent in a \textit{formal proof}. \\

% Tree Structure of Proofs
\begin{center}
	\begin{tikzpicture}[scale=0.2]
	
	\coordinate (L) at (135:10);
	\coordinate (R) at (45:10);
	
	% Level 0
	\draw [black] (0:0) -- ++(L);
	\draw [black] (0:0) -- ++(R);
	
	\fill [white] (0:0) circle (3);
	\draw [black] (0:0) circle (3);
	\draw (0:0) node {$F$};
	
	% Level 1
	\draw [black] (R) -- ++(L);
	\draw [black] (R) -- ++(R);
	
	\fill [white] (L) circle (3);
	\draw [black] (L) circle (3);
	\draw (L) node {$P_1$};
	
	\fill [white] (R) circle (3);
	\draw [black] (R) circle (3);
	\draw (R) node {$I_2$};
	
	% Level 2
	\draw [black] ($ 2*(R) $) -- ++(L);
	\draw [black] ($ 2*(R) $) -- ++(R);
	
	\fill [white] ($ (R) + (L) $) circle (3);
	\draw [black] ($ (R) + (L) $) circle (3);
	\draw ($ (R) + (L) $) node {$P_2$};
	
	\fill [white] ($ 2*(R) $) circle (3);
	\draw [black] ($ 2*(R) $) circle (3);
	\draw ($ 2*(R) $) node {$I_1$};
	
	% Level 3
	\fill [white] ($ 2*(R) + (L) $) circle (3);
	\draw [black] ($ 2*(R) + (L) $) circle (3);
	\draw ($ 2*(R) + (L) $) node {$P_3$};
	
	\fill [white] ($ 3*(R) $) circle (3);
	\draw [black] ($ 3*(R) $) circle (3);
	\draw ($ 3*(R) $) node {$P_4$};
	
	% Arrows	
	\foreach \i in {0,1,2} {
		\coordinate (A) at ($ (45:3) + \i*(R) $);
		\coordinate (B) at ($ (135:3) + \i*(R) $);
		\fill [black,rotate around={45:(A)}] (A) -- ++(0.8,0.5) -- ++(0,-1);
		\fill [black,rotate around={45:(B)}] (B) -- ++(-0.5,0.8) -- ++(1,0);
	}
	
	\end{tikzpicture}
\end{center}
\vspace{4mm}

In this \textit{tree}, the \textit{first premises} are labeled as $P_n$ where $n\in\{1,2,3,4\}$, the \textit{intermediate conclusions} are labeled as $I_m$ where $m\in\{1,2\}$, and the \textit{final conclusion} is labeled $F$. Though one can make a distinction between first premises and intermediate conclusions, it is important to note that they are all premises with regard to the eventual final conclusion. Thus, there are no restrictions on combining them beyond those dictated by the rules of inference. \\

In modern proof theory, proofs like these are studied not only for their semantic meaning, but also for their mathematical structure. They are treated like \textit{data structures}, which are abstract objects that are used in computer science to model the storage and flow of digital \textit{information}. Similarly, a proof is an abstract text that models the flow of logical information as it is manipulated by rules of inference toward a final conclusion. \\

Information, in the context of computer science and, more generally, \textit{information theory}, refers to a \textit{syntactic} message. Formally, it is an order sequence of symbols \\

% Data

In computer science, the data structures of interest are \textit{recursive}. That is, the structure can be defined "in terms of itself." \textit{Recursion} is a deep topic, and we will spend much of the next part characterizing it, but a practical way to think about a recursive structure is in terms of \textit{type}. \\

Types are a widespread feature in programming languages. As a concept, they are founded in a logical system called a \textit{type theory}. There are multiple type theories, and some of them are used as alternative foundations for mathematics. We will cover all of this in detail later, but, essentially, a type is a label that can be given to 

% Proofs and models
% String together arguments to make proofs
% Tree structure of arguments
% List structure of arguments
% Data structures structure data in such a way that it can be interpreted as a model of semantic ideas.
% Proofs structure text in such a way that...
% Information is syntactic, data is semantic

However, not all forms of reasoning have such strict rules. There are other methods of reasoning that cannot be modeled by an argument. In contrast to logical reasoning, \textit{intuitive reasoning} has steps that are \textit{not} understood. Although the question might seem peculiar, it is worth asking whether or not computation can be intuitive. So, to begin our journey of understanding computation in a modern, logical sense, we will first walk in the other direction, considering it instead in a mystical, otherworldly sense. \\

%----------------------------------------

\subsection{The Oracle, the Seer, and the Sage}

\textit{Intuition} is the capacity to create conclusions without evidence, proof, or a combination of the two. If any premises are involved, they may appear to an onlooker as if they were plucked out of thin air. If any method is involved, it is esoteric or hidden from sight. Acts of intuition range from the mundane procedures we perform without thinking to great feats of intellectual, artistic, and athletic achievement. And while it is often associated with magic or supernatural ability, intuition is a real, observable phenomenon, and it is a form of reasoning. \\

Like that of reasoning, the definition of intuition is fuzzy. There are a variety of events that one might label as the product of intuition that are actually quite functionally distinct from one another. For this reason, I would like to consider and compare three archetypes that are known for their intuitive skills: the Oracle, the Seer, and the Sage. For the skeptics among you, I ask that you suspend your disbelief for a moment and assume that our characters are acting in good faith. There is no lying here; the conclusions are sincere. \\

\paragraph{The Oracle} \hspace*{1mm} \vspace*{2mm}

An Oracle is a person who predicts the future by acting as a vessel for a god or a set of gods. They are considered by believers to be portals through which the divine speaks directly. They are found in histories all over the world, but most people associate the role with the priestesses of Ancient Greece. Picture a woman with eyes that roll back into her head as she speaks in a possessed, thundering voice. \\

For an Oracle, the conclusions come straight from the source. She may not even do any reasoning herself, save for the \textit{unconscious reasoning} that is performed while she is possessed. However, conscious or not, she provides a great service to mankind. There is no clearer answer than the one given to you directly from the gods themselves, even if it has to be sent through what is essentially the human version of a telephone. \\

For those seeking counsel, the premises of the Oracle's conclusion are unnecessary because they just \textit{know} that the statement must be true. For them, the connection the Oracle has with the gods and with reality is part of life itself. We all have deep beliefs like this. For example, most people do not feel that gravity needs to be proven to them in order for them to accept it as fact. Their perception of gravity in the world around them is proof enough. This is the kind of intuition characterized by \textit{subconscious reasoning}, the reasoning you do without being aware of doing it. \\

\paragraph{The Seer} \hspace*{1mm} \vspace*{2mm}

A Seer is a person who predicts the future by interpreting signs from a god or a set of gods. Unlike an Oracle, a Seer speaks divine truth in his own words, drawing from an innate, sometimes god-given power to see meaning in natural events or occult objects. There are various methods of divination that are used by Seers, such as scrying (the gazing into magical things, like crystal balls, for the purpose of seeing visions), auspicy (the interpretation of bird migration), or dowsing (the use of magic to find water, often with the aid of a dowsing rod). The acceptance of any particular method is cultural, but beliefs in divination vary greatly, even within a single society. \\

The Seer has premises for his conclusion, but the rules of inference involved in his reasoning are incomprehensible to others. He 

\paragraph{The Sage} \hspace*{1mm} \vspace*{2mm}

% Seers have premises and conclusions, oracles just speak conclusions
% Does intuition involve unconcious, preconcious, or concious reasoning?
% Is intuition computation?
% This book is as much about intuition as it is about computation.
% In some sense, intuition is thought of as the opposite of logic (it has no argument or basis in reality)
% In another sense, logical truths (tautologies) could be considered intuitive, so all of logic could be the product of intuition. Unless you believe that logic is objective.
% The Oracle and the Seer are unique from the sage because they KNOW their truth is absolute.
% Computers are honestly closer to oracles or seers than sages.
% Buddhism - intuition is a faculty of the mind of immediate knowledge

\begin{tcolorbox}[breakable, enhanced, colback=textbook-blue, sharp corners]
	\vspace{3mm}
	\begin{center}
		\textbf{What is Truth?}
	\end{center}
	Truth, in the absolute sense, has been discussed and debated since the dawn of man. It is a concept that seems obvious to us, and yet it always seems to elude our understanding. Philosophers have formulated dozens of theories of truth over the years, with some asserting that truth is an objective property of our universe and others asserting that truth is a useful lie that mankind has invented. And while there is merit in those claims that truth is not real, we still see around us the technology that was born out of our intuitive sense of true and false. Especially in computer science where everything is represented in binary. \\
	
	The traditional theories of truth have been termed \textit{substantive}. That is, they assert that truth has some basis in reality and that it is a meaningful thing to discuss. Early theories of truth from Ancient Greece are considered the foundation of \textit{correspondence theory}, the idea that statements are true if their symbols are arranged in such a way that they express an abstract thought that accurately describes reality. This theory defines truth in the context of a relationship between language and objective reality. It is a useful philosophy that allows us to orient ourselves in the world around us. However, many philosophers believe that truth cannot be explained by such a simple rule. In fact, there are many more factors that could play a role in the concept. \\
	
	Objections to a strict correspondence theory usually take issue with the treatment of language as something monolithic and easy to classify within a true-false dichotomy. For example, a statement encoded as a sentence in a language is only meaningful to people who can read that language. What does this imply about a particular language's relationship with truth? Is a statement only true for those who understand it? Furthermore, there is no guarantee that readers of the same language will even be able to agree on the meaning expressed by a particular sentence. Languages often encompass multiple dialects that might parse words differently. Words themselves also may not be precise, and some abstract thoughts may not have suitable words in some languages. These are the points raised in \textit{social constructivism}, a theory which avers that human knowledge is historically and culturally specific. Social constructivists also believe that truth is \textit{constructed} and that language cannot faithfully represent any kind of objective reality. \\
	
	Other substantive theories find the essence of truth nestled in other abstract concepts. \textit{Consensus theory}, as the name implies, defines truth as something that is agreed upon either by all human beings or by a subset (such as academic groups or standards bodies). This is another anthropocentric definition that is at constant risk of philosophical division on any given topic. \textit{Coherence theory} takes a more objective approach, claiming that a statement can only be true if it fits into a system of statements that support each other's truth. This is similar to the notion of \textit{formal systems}, which we will discuss in depth later. However, traditional coherence theories attempt to explain all of reality within a single coherent system of truths, which is incompatible with our modern understanding of formal systems. \\
	
	Modern developments in philosophy have resulted in theories that deviate from the long-held, substantive opinions on the nature of truth. These \textit{minimalist} theories assert instead that truth is either not real or not a necessary property of language or logic. They claim that statements are \textit{assertions} and thus are implicitly understood to be true. For example, it is understood that by putting forth the sentence "$2+2=4$," you are endorsing the semantic meaning "$2+2=4$ is true." The clause "is true" is called a \textit{truth predicate}, and minimalist theories of truth often consider its use redundant. \\
	
	This idea of a truth predicate is borrowed from Alfred Tarski's \textit{semantic theory of truth}, which is a substantive theory that refines the correspondence concepts espoused by Socrates, Plato, and Aristotle for formal languages. This theory makes a distinction between a statement made in a formal language and a truth predicate, which evaluates the truth of the statement. Tarski made this distinction in order to circumvent the \textit{liar's paradox}, which is often presented with the following example: "This sentence is false." If the predicate "is false" is considered to be part of "this sentence," the truth of this statement cannot be decided. For this reason, Tarski states that a language cannot contain its own truth predicate. This is enforced by requiring that "this sentence" be written in an \textit{object language} and that "is false" be written in a \textit{metalanguage}. \textit{Convention T}, the general form of this rule, can be expressed as
	
	\begin{center}
		\textit{"$P$" is true if and only if $P$}
	\end{center}
	
	where "$P$" is simply the metalanguage sentence $P$ rendered in the object language. That is, the \textit{syntactic representation} is assigned a \textit{truth value} of true if and only if the semantic meaning it represents is considered true (according to whatever theory of truth you employ). By this rule, we say that "$2+2=4$" is true if and only if \textit{the sum of the number $2$ with the number $2$ is equal to the number $4$}. Minimalist redundancy theories modify Tarski's theory by interpreting "$P$" as an implicit assertion of the truth of $P$. The sentence "$P$," which asserts that $P$ is true, is then false if and only if its semantic meaning $P$ is false. \\
	
	\parbreak
	\vspace{\baselineskip}
	
	The debate on the nature of truth rages on \textit{in perpetuum}. However, for our purposes, we must be practical. There is truth in logic and mathematics and computer science as well, but, in practice, it has little to do with the debates on \textit{absolute truth} described above. Their truth is \textit{relative}. That is, it relies on the assumption that our premises are true. Cognizant of this, we move forward with our thinking, searching for truths within these arbitrary boundaries. \\
	\vspace{3mm}
\end{tcolorbox}
\vspace{2\baselineskip}

%----------------------------------------

\subsection{The Mathematician and the Scientist}

% Deductive and Inductive (+ Abductive)
% Mathematician write proofs (deductive), but they also use creativity (inductive)
% Is creativity inductive?
% Human computer

Logical reasoning is the \textit{inference} of new information from present information. It involves \textit{rule of inferences} that are used to relate sets of \textit{premises} to \textit{conclusions}. There are three kinds of logical reasoning: \textit{deductive}, \textit{inductive}, and \textit{abductive}. Each is classified according to which piece of information (premises, rule, or conclusion) is missing and must be inferred from the others. \\

% Explanatory vs ampliative
% Certainty vs uncertainty (probability)
% Top-down vs bottom-up
% Logical truths (tautologies) vs facts
% Abduction = induction + justification
% Closed and open world assumptions
% Domain of discourse

Inductive arguments exist on a spectrum between \textit{weak} and \textit{strong}. Those that are stronger and more persuasive have a higher probability of having a true conclusion. Inductive reasoning is associated with \textit{science} and \textit{critical thinking} because it allows one to make generalizations about complex phenomena given limited evidence. Unlike deductive reasoning, it attempts to find new knowledge that is not simply contained within its premises. \\

Statements made by induction are bolstered with evidence whereas deductive statements are as true as their premises. This leaves inductive reasoning susceptible to \textit{faulty generalizations} and \textit{biased sampling}. Induction must also assume that future events will occur exactly as they have in the past, which is not always the case. For example, a turkey that is fed every morning with the ring of a bell may infer by induction that bell $\rightarrow$ food. However, he will see the error in his reasoning when the farmer rings the bell on Thanksgiving Day and instead slits his throat. \\

\paragraph{Abductive Reasoning} \hspace*{1mm} \vspace*{2mm}

\textit{Abductive reasoning} is the inference of a \textit{premise}, given a conclusion and a rule. It is investigative in nature. For example, given a conclusion "The grass is wet" and a rule "When it rains, the grass gets wet," we might determine that rain is the best explanation for the wetness of the grass. Thus, we abduce that "It might have rained." \\

Like induction, abduction can also produce a hypothesis. However, abduction does not seek a new relationship between two previously unconnected statements. Rather, it uses established relationships to find a reasonable explanation for a statement that is assumed to be true. It is often used by detectives or diagnosticians who need to find a probable cause of an event. It is also used in Bayesian statistics. While multiple premises may be abduced, typically we want to abduce a single, "best" premise. \\

Abductive reasoning allows us to ignore the many causes that are unlikely in favor of those few that may be relevant to the problem at hand. For example, doctors are often taught to heed the following proverb: "When you hear hoofbeats, think of horses, not zebras." That is, when a patient exhibits certain symptoms, a doctor should abduce from them a commonplace disease before considering more exotic possibilities. However, "zebras" do exist. Sometimes the most likely cause is not the actual cause. For this reason, abduction is also considered to be equivalent to a deductive fallacy called \textit{affirming the consequent}, which is like a \textit{modus ponens} performed in reverse. That is, given a conditional $P\rightarrow Q$ and the consequent $Q$, abduction infers $P$ from $Q$ by assuming that the converse $Q\rightarrow P$ of the conditional is also true. This is not a deductively valid inference. Considering our example again, the grass might be wet from rain, but this is not \textit{necessarily} true. It is also possible that the sprinkler system is on. Or perhaps there is a zebra-esque scenario like a flood. \\\\

%-------

\begin{tcolorbox}[breakable, enhanced, colback=textbook-blue, sharp corners]
	\vspace{3mm}
	\begin{center}
		\textbf{Induction and Abduction}
	\end{center}
	% Can computers induce?
	% Machine learning simulates induction deductively
	% Deductive computers, AI, General AI, Artificial Conciousness
	% Calculation vs Computation
	
	% Fuzzy expert systems "abduce"
	TEXT
	\vspace{3mm}
\end{tcolorbox}
\vspace{2\baselineskip}

%----------------------------------------

\subsection{Mechanical Computation}

% Computation done by humans mechanically with tools
% Computation done entirely by machines

Before we discuss and classify automata in depth, we should first consider what is \textbf{not} an automaton. What is an example of something that might perform some kind of calculation, but is not a computer? What about a microwave? Is a microwave a computer? No, it is not. A computer can be programmed in some meaningful, robust way. A microwave contains a microprocessor, which uses \textit{combinational logic} and basic binary inputs to set timers and operate the oven. It cannot be programmed in any meaningful way. What about a calculator? Is a calculator a computer? If we are being formal, the answer is no, but it depends on what kind of "calculator" we are talking about. \\

Calculators, such as \textit{counting boards} and the \textit{abacus}, have been around since pre-history. Calculators with four-function capabilities have been around since the invention of Wilhelm Schickard's mechanical calculator in 1642. In the late 19th century, the \textit{arithmometer} and comptometers, two kinds of digital, mechanical calculator, were being used in offices. The Dalton Adding Machine introduced buttons to the calculator in 1902, and pocket mechanical calculators became widespread in 1948 with the \textit{Curta}. None of these are computers. \\

The difference between a calculator and a computer is that a computer can be programmed and a calculator cannot. What does it mean to be programmable? That is perhaps the central question of automata theory, and we will discuss in this section several levels of "programmability." However, for now, we can certainly say that a simple, four-function electronic calculator is not a computer. It simply uses combinational circuits like full-adders, full-subtractors, multipliers, and dividers to implement its functions, and there is no potential for modifications or user-defined functions. \\

Surprisingly enough, \textit{special-purpose computers} have also been around for a long time. Early examples include the \textit{Antikythera mechanism} (an Ancient Greek analog computer), Al-Jazari's 1206 water-powered \textit{astronomical clock}, the 1774 \hrefcolor[blue]{https://www.youtube.com/watch?v=bY\_wfKVjuJM}{\underline{Writer Automaton}} (a mechanical doll that could be programmed to write messages), the 1801 \textit{Jacquard loom}, and \textit{slide rules}. Some later mechanical computers were quite powerful, such as \textit{differential analyzers} (special-purpose computers for solving differential equations) and fire control computers. Charles Babbage designed the \textit{Analytical Engine}, a general-purpose computer, in 1837, and Ada Lovelace wrote a program for it, but it was never built. General-purpose computing would not reemerge until the 1940s. \\

The line between calculator and computer began to blur with the introduction of \textit{programmable calculators} in the 1960s. Many modern high-end calculators are programmable in Turing-complete languages such as TI-BASIC or even C or C++, which officially makes them computers. Once we start implementing \textit{sequential logic} with components like SR latches or D flip-flops, we are storing state, and state is a requirement for computing. Circuits that use sequential logic can be considered automata and, given enough complexity, computers. \\

\begin{tcolorbox}[breakable, enhanced, colback=textbook-blue, sharp corners]
	\vspace{3mm}
	\begin{center}
		\textbf{Modern Computing is American}
	\end{center}
	TEXT
	\vspace{3mm}
\end{tcolorbox}
\vspace{2\baselineskip}

% Hey, let's build machines to do this instead.
% Wait, what exactly are we building? Let's model the problem.
% Computation as a mechanical process vs a function.
% Abstract away the mechanics, and a function becomes a conversation (input and output).
% Computing is more than just math, semantically.
% Structure of language, types of sentences (declarative/imperative)
